---
title: "Getting Started with ADS 8192"
author: "Jared Andrews"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with ADS 8192}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  eval = FALSE
)
```

## Overview

**ADS 8192: Developing Scientific Applications** teaches graduate students to
build scientific software in R using established software engineering principles
вҖ” separation of concerns, DRY (Don't Repeat Yourself), composability, and
layered architecture. This package provides the reference implementation: a set
of PCA analysis functions for SummarizedExperiment objects, a Shiny interactive
explorer, and a command-line interface.

## Installation

```{r install, eval=FALSE}
# Install Bioconductor dependencies first
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install(c("SummarizedExperiment", "airway"))

# Install the course package
remotes::install_github("St-Jude-MS-ABDS/ADS8192")
```

## Quick Start

```{r quick-start}
library(ADS8192)
library(airway)

# Load the airway dataset (RNA-seq, dexamethasone treatment vs. control)
data("airway", package = "airway")
airway
```

### Run PCA

```{r run-pca}
result <- run_pca(airway, n_top = 500)

# View the PCA scores merged with sample metadata
head(result$scores)
```

### Visualize

```{r plot-pca}
plot_pca(result, color_by = "dex")
```

```{r plot-pca-both}
plot_pca(result, color_by = "dex", shape_by = "cell")
```

### Check Variance Explained

```{r variance}
pca_variance_explained(result)
```

## Architecture: Separation of Concerns

The package is organized around a principle that shows up in virtually every
well-engineered codebase: **separate _what_ you compute from _how_ users
access it**. The analysis logic lives in small, testable, composable R functions
(the core layer). Presentation layers вҖ” an R API, a Shiny app, and a CLI вҖ” are
thin wrappers that delegate to those same core functions.

```
             Core Analysis Functions
  run_pca() вҶ’ plot_pca() вҶ’ save_pca_results()
        вҶ‘            вҶ‘            вҶ‘
   в”Ңв”Җв”Җв”Җв”Җв”ҙв”Җв”Җв”Җв”Җв”җ  в”Ңв”Җв”Җв”Җв”Җв”ҙв”Җв”Җв”Җв”Җв”җ  в”Ңв”Җв”Җв”Җв”ҙв”Җв”Җв”Җв”Җв”Җв”җ
   в”Ӯ R API   в”Ӯ  в”Ӯ Shiny   в”Ӯ  в”Ӯ  CLI    в”Ӯ
   в”Ӯ (users) в”Ӯ  в”Ӯ (web)   в”Ӯ  в”Ӯ(scripts)в”Ӯ
   в””в”Җв”Җв”Җв”Җв”Җв”Җв”Җв”Җв”Җв”ҳ  в””в”Җв”Җв”Җв”Җв”Җв”Җв”Җв”Җв”Җв”ҳ  в””в”Җв”Җв”Җв”Җв”Җв”Җв”Җв”Җв”Җв”ҳ
```

Because every interface calls the **same core functions**, you get several
properties for free:

- **DRY (Don't Repeat Yourself)** вҖ” fix a bug or add a feature once, and every
  interface benefits immediately.
- **Testability** вҖ” core functions are pure R with well-defined inputs and
  outputs, so unit tests cover the real computation regardless of how it's
  invoked.
- **Composability** вҖ” each function does one thing, so users (and other
  packages) can recombine them in ways you didn't anticipate.

This layered design is why the package вҖ” and the course lectures вҖ” are
structured the way they are: we build the core functions first (Lecture 03),
package and test them (Lectures 05вҖ“06), then add presentation layers one at a
time (Shiny in 07вҖ“08, CLI in 09вҖ“10).

### R API

```{r r-api, eval=FALSE}
library(ADS8192)
library(airway)
data("airway", package = "airway")
result <- run_pca(airway, n_top = 500)
plot_pca(result, color_by = "dex")
```

### Shiny App

```{r shiny, eval=FALSE}
ADS8192::run_app()
```

### CLI (via Rapp)

```bash
ADS8192 pca --counts counts.tsv --meta samples.tsv --output results/
```

## Course Lectures

The package includes all lecture materials as pkgdown articles. See the
"Course Materials" dropdown in the navigation bar.
