---
title: "Getting Started with ADS 8192"
author: "Jared Andrews"
date: sys.Date()
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with ADS 8192}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Overview

**ADS 8192: Developing Scientific Applications** teaches the "three interfaces,
one core" architecture for scientific software in R. This package provides the
reference implementation: a set of PCA analysis functions for
SummarizedExperiment objects, a Shiny interactive explorer, and a command-line
interface.

## Installation

```{r install, eval=FALSE}
# Install Bioconductor dependencies first
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install(c("SummarizedExperiment", "airway"))

# Install the course package
remotes::install_github("YOUR-USERNAME/ADS8192")
```

## Quick Start

```{r quick-start}
library(ADS8192)

# Load the example dataset (100 genes, 8 samples)
data(example_se)
example_se
```

### Run PCA

```{r run-pca}
result <- run_pca(example_se, n_top = 50)

# View the PCA scores merged with sample metadata
head(result$scores)
```

### Visualize

```{r plot-pca}
plot_pca(result, color_by = "treatment")
```

```{r plot-pca-both}
plot_pca(result, color_by = "treatment", shape_by = "batch")
```

### Check Variance Explained

```{r variance}
pca_variance_explained(result)
```

## The "Three Interfaces, One Core" Architecture

```
                Package Core
  make_se() → run_pca() → plot_pca()
        ↑            ↑            ↑
   ┌────┴────┐  ┌────┴────┐  ┌───┴─────┐
   │ R API   │  │ Shiny   │  │  CLI    │
   │ (users) │  │ (web)   │  │(scripts)│
   └─────────┘  └─────────┘  └─────────┘
```

All three interfaces call the **same core functions**. Fix a bug once, and it's
fixed everywhere.

### Interface 1: R API

```{r r-api, eval=FALSE}
library(ADS8192)
result <- run_pca(se, n_top = 500)
plot_pca(result, color_by = "treatment")
```

### Interface 2: Shiny App

```{r shiny, eval=FALSE}
ADS8192::run_app()
```

### Interface 3: CLI (via Rapp)

```bash
ADS8192 pca --counts counts.tsv --meta samples.tsv --output results/
```

## Course Lectures

The package includes all lecture materials as pkgdown articles. See the
"Course Materials" dropdown in the navigation bar.
