---
title: "Course Setup"
author: "Jared Andrews"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Course Setup}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

## Overview

This guide installs **every** R package you may need for ADS 8192 — the shared
development stack, the course reference package, and the dataset/analysis
packages for all 13 homework projects. Running the code block below once on a
fresh R installation will get you fully set up.

> **Tip:** If you have already installed some of these packages, re-running
> `install.packages()` or `BiocManager::install()` will simply skip packages
> that are already up to date.

## Prerequisites

- **R ≥ 4.5.0** — install from [CRAN](https://cran.r-project.org/)
- **RStudio** (recommended) or Positron

## Install All Packages

Copy and run the entire block below in your R console. It installs:
1. BiocManager (needed to install Bioconductor packages)
2. CRAN packages (development tools, Shiny stack, analysis dependencies)
3. Bioconductor packages (data containers, datasets, annotation tools)

```{r install-all}
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")

cran_pkgs <- c(
  # Development & documentation
  "devtools", "usethis", "roxygen2", "testthat", "pkgdown",
  "knitr", "rmarkdown", "remotes",
  # Shiny
  "shiny", "bslib", "DT",
  # Core
  "ggplot2", "rlang",
  # CLI
  "Rapp",
  # Project-specific
  "uwot",
  "cluster",
  "msigdbr",
  "igraph"
)

install.packages(cran_pkgs)

bioc_pkgs <- c(
  # Data containers
  "SummarizedExperiment",
  "SingleCellExperiment",
  # Analysis
  "DESeq2",
  # Visualization
  "ComplexHeatmap",
  "dittoSeq",
  # Datasets
  "airway",
  "macrophage",
  "scRNAseq",
  "tximeta",
  "recount3",
  "curatedTCGAData",
  "TCGAutils",
  "TENxPBMCData"
)

BiocManager::install(bioc_pkgs)
```

## Verify Installation

After installation, run the following to confirm the key packages load without
error:

```{r verify}
library(devtools)
library(SummarizedExperiment)
library(SingleCellExperiment)
library(ggplot2)
library(shiny)
library(testthat)
```

If any package fails to load, re-run the relevant `install.packages()` or
`BiocManager::install()` call for that package and check the error message.

## Install the Course Reference Package

The ADS8192 reference package can be installed directly from GitHub:

```{r install-ads8192}
remotes::install_github("St-Jude-MS-ABDS/ADS8192")
```

## Session Info

```{r session-info}
sessionInfo()
```
