---
title: "Course Setup"
author: "ADS 8192"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Course Setup}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

## Overview

This guide installs **every** R package you may need for ADS 8192 — the shared
development stack, the course reference package, and the dataset/analysis
packages for all 13 homework projects. Running the code block below once on a
fresh R installation will get you fully set up.

> **Tip:** If you have already installed some of these packages, re-running
> `install.packages()` or `BiocManager::install()` will simply skip packages
> that are already up to date.

## Prerequisites

- **R ≥ 4.1.0** — install from [CRAN](https://cran.r-project.org/) or use
  your institution's R distribution.
- **RStudio** (recommended) or any editor with R terminal support.
- A working internet connection for downloading packages.

## Install All Packages

Copy and run the entire block below in your R console. It installs:
1. BiocManager (needed to install Bioconductor packages)
2. CRAN packages (development tools, Shiny stack, analysis dependencies)
3. Bioconductor packages (data containers, datasets, annotation tools)

```{r install-all}
# --- 1. BiocManager (Bioconductor installer) --------------------------------
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")

# --- 2. CRAN packages -------------------------------------------------------
# Development & documentation tools
#   devtools, usethis, roxygen2, testthat, pkgdown, knitr, rmarkdown, remotes
# Shiny stack
#   shiny, bslib, DT
# Core analysis
#   ggplot2, rlang
# CLI
#   Rapp
# Project-specific (CRAN)
#   uwot        (Project 1 — UMAP)
#   cluster     (Project 4 — K-means)
#   msigdbr     (Project 7 — gene set scoring)
#   igraph      (Project 9 — gene correlation network)

cran_pkgs <- c(
  # Development & documentation
  "devtools", "usethis", "roxygen2", "testthat", "pkgdown",
  "knitr", "rmarkdown", "remotes",
  # Shiny
  "shiny", "bslib", "DT",
  # Core
  "ggplot2", "rlang",
  # CLI
  "Rapp",
  # Project-specific
  "uwot",
  "cluster",
  "msigdbr",
  "igraph"
)

install.packages(cran_pkgs)

# --- 3. Bioconductor packages -----------------------------------------------
# Data containers
#   SummarizedExperiment, SingleCellExperiment
# Visualization
#   ComplexHeatmap
# Datasets (one per project)
#   airway           (Project 0  — PCA Explorer, reference)
#   scRNAseq         (Projects 1, 4, 5, 6, 8, 12, 13)
#   parathyroidSE    (Project 2  — Sample Similarity)
#   recount3         (Projects 3, 9 — Diff. Expression, Gene Corr Network)
#   curatedTCGAData  (Projects 7, 10 — Gene Set Scoring, Expression Heatmap)
#   TCGAutils        (Projects 7, 10 — helper for curatedTCGAData)
#   TENxPBMCData     (Project 11 — Dimensionality Estimation)

bioc_pkgs <- c(
  # Data containers
  "SummarizedExperiment",
  "SingleCellExperiment",
  # Visualization
  "ComplexHeatmap",
  # Datasets
  "airway",
  "scRNAseq",
  "parathyroidSE",
  "recount3",
  "curatedTCGAData",
  "TCGAutils",
  "TENxPBMCData"
)

BiocManager::install(bioc_pkgs)
```

## Verify Installation

After installation, run the following to confirm the key packages load without
error:

```{r verify}
library(devtools)
library(SummarizedExperiment)
library(SingleCellExperiment)
library(ggplot2)
library(shiny)
library(testthat)
```

If any package fails to load, re-run the relevant `install.packages()` or
`BiocManager::install()` call for that package and check the error message.

## Install the Course Reference Package

The ADS8192 reference package can be installed directly from GitHub:

```{r install-ads8192}
remotes::install_github("St-Jude-MS-ABDS/ADS8192")
```

## Session Info

```{r session-info}
sessionInfo()
```
