---
title: "Lecture 3: Data Structures & R Ecosystems (Bioconductor)"
author: "Jared Andrews"
date: "April 14, 2026"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  eval = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 6
)
```

# Learning Objectives

By the end of this session, you will be able to:

1. Generate a `SummarizedExperiment` object from raw assay data and sample metadata
2. Find and use Bioconductor package documentation (vignettes/manuals) to select appropriate tools
3. Visualize elements of a `SummarizedExperiment` (assays, rowData, colData) with a Bioconductor visualization package
4. Extract assay data from a `SummarizedExperiment` and use it to run PCA
5. Write small, testable functions that take structured objects as input and return well-defined outputs

**Course Learning Outcomes (CLOs):** CLO 1, 3

---

# Why Structured Data Containers?

## The Problem with Loose Data

Imagine you're analyzing gene expression data. You might have:

- A **counts matrix**: genes × samples
- A **sample metadata table**: sample info (treatment, batch, etc.)
- A **gene annotation table**: gene symbols, descriptions, etc.

If these are three separate objects, you face several problems:

```{r bad-example, eval=FALSE}
# Dangerous: Three separate objects that can get out of sync
counts <- read.csv("counts.csv", row.names = 1)
sample_info <- read.csv("sample_metadata.csv")
gene_info <- read.csv("gene_annotations.csv")

# What happens if you subset counts but forget to subset sample_info?
counts_subset <- counts[, 1:5]
# sample_info still has all samples - now mismatched!

# What if you accidentally reorder one but not the others?
counts <- counts[, order(colnames(counts))]
# Now sample_info doesn't match the column order!
```

## The Solution: Structured Containers

Bioconductor's `SummarizedExperiment` class keeps everything synchronized:

```
                    colData (sample metadata)
                    ┌───────────────────────┐
                    │ sample treatment batch │
                    │ S1     trt      A     │
                    │ S2     ctrl     A     │
                    │ S3     trt      B     │
                    └───────────────────────┘
                              ↓
              ┌───────────────────────────────┐
              │        assays (counts)         │
   rowData    │         S1    S2    S3         │
   (genes)  → │ gene1  100   150   120         │
              │ gene2   50    75    60         │
              │ gene3  200   180   210         │
              └───────────────────────────────┘
```

When you subset or reorder a `SummarizedExperiment`, all components stay synchronized automatically!

---

# Group Discussion + Pseudocode: Designing a Gene Data Structure

**Prompt:** "How would you create a data structure to represent a gene?"

Discuss in groups and sketch a simple structure. Think about:

- Which fields are **core** (required) vs **derived** (can be computed)?
- Which fields are **optional** vs **always present**?
- How will you ensure **validity** (e.g., start <= end, strand in {+,-})?

### Simple Pseudocode (List-Based)

```r
gene <- list(
    gene_id = "ENSG00000141510",
    symbol = "TP53",
    chr = "17",
    start = 7661779,
    end = 7687550,
    strand = "-",
    biotype = "protein_coding",
    length = 7687550 - 7661779 + 1  # derived
)
```

### S4-Style Sketch (Class + Constructor)

```r
setClass("Gene",
         slots = c(
             gene_id = "character",
             symbol = "character",
             chr = "character",
             start = "integer",
             end = "integer",
             strand = "character",
             biotype = "character"
         ))

Gene <- function(gene_id, symbol, chr, start, end, strand, biotype) {
    new("Gene",
        gene_id = gene_id,
        symbol = symbol,
        chr = chr,
        start = as.integer(start),
        end = as.integer(end),
        strand = strand,
        biotype = biotype
    )
}
```

---

## Why S4 in Bioconductor?

Bioconductor packages rely heavily on **S4 classes** because they provide:

- **Formal class definitions** (clear structure and types)
- **Validity checks** (catch invalid objects early)
- **Method dispatch** (functions behave differently based on object class)

`SummarizedExperiment` is an S4 class. Its formal structure and methods keep assays, rowData, and colData synchronized and enforce consistency.

## S4 Best Practices (Design Checklist)

- Keep **slots minimal** and focused on essential fields
- Write **validity checks** for invariants (e.g., `start <= end`)
- Provide **constructor functions** so users don't call `new()` directly
- Use **accessor methods** instead of direct slot access (`@`)
- Document **invariants** and expected slot types
- Avoid storing **redundant derived values** when they can be computed

---

# Setup: Installing Required Packages

If you haven't already, install the required Bioconductor packages:

```{r install-packages, eval=FALSE}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install(c(
    "SummarizedExperiment",
    "airway",
    "ComplexHeatmap"
))

# Also install ggplot2 for visualization
install.packages("ggplot2")
```

Now load the libraries:

```{r load-libraries}
library(SummarizedExperiment)
library(airway)
library(ComplexHeatmap)
library(ggplot2)
```

---

# The `airway` Dataset

We'll use the `airway` dataset, which contains RNA-seq data from airway smooth muscle cells. The experiment compared cells treated with dexamethasone (a corticosteroid) to untreated cells.

```{r load-airway}
data("airway")
airway
```

## Exploring the Structure

### What class is this object?

```{r class}
class(airway)
```

`RangedSummarizedExperiment` is a subclass of `SummarizedExperiment` that also stores genomic ranges for each feature (gene).

### Dimensions: How many genes and samples?

```{r dimensions}
dim(airway)
nrow(airway)  # genes (features)
ncol(airway)  # samples
```

### Sample Metadata (colData)

The `colData()` function extracts sample-level metadata:

```{r coldata}
colData(airway)
```

Key variables:
- `cell`: Cell line identifier
- `dex`: Treatment status - "trt" (dexamethasone) or "untrt" (control)

```{r coldata-access}
# Access as a data.frame
as.data.frame(colData(airway))

# Access specific columns
airway$dex
airway$cell
```

### Gene/Feature Data (rowData)

The `rowData()` function extracts feature-level (gene) metadata:
                       
```{r rowdata}
rowData(airway)
```

### Assay Data (the counts!)

The actual expression data is stored in "assays". This dataset has one assay called "counts":

```{r assays}
# What assays are available?
assayNames(airway)

# Get the counts matrix
counts_matrix <- assay(airway, "counts")
# Or equivalently: assays(airway)$counts

# Check dimensions
dim(counts_matrix)

# Preview first few genes and samples
counts_matrix[1:5, 1:4]
```

> **Exercise A (Warm-up):** Use `head()` and `tail()` to explore the counts matrix. What is the range of count values? Use `summary()` on a single column to find out.

---
 
# Building the Analysis Core

Now we'll build a set of reusable functions that form the "analysis core" of our package. These functions will later be:

1. Packaged into an R package (Lecture 5)
2. Wrapped by a Shiny app (Lectures 7-8)
3. Exposed via a CLI (Lectures 9-10)

## Function 1: `make_se()` — Create a SummarizedExperiment

This function creates a `SummarizedExperiment` from a counts matrix and metadata data.frame:

```{r make-se}
#' Create a SummarizedExperiment from counts and metadata
#'
#' @param counts A matrix of counts (genes × samples). Row names should be 
#'   gene identifiers, column names should be sample identifiers.
#' @param col_data A data.frame of sample metadata. Row names must match 
#'   column names of counts.
#' @param row_data Optional data.frame of gene metadata. Row names must match 
#'   row names of counts.
#'
#' @return A SummarizedExperiment object
#' 
#' @examples
#' counts <- matrix(rpois(100, 50), nrow = 10, ncol = 10)
#' rownames(counts) <- paste0("gene", 1:10)
#' colnames(counts) <- paste0("sample", 1:10)
#' meta <- data.frame(
#'   treatment = rep(c("ctrl", "trt"), each = 5),
#'   row.names = colnames(counts)
#' )
#' se <- make_se(counts, meta)
make_se <- function(counts, col_data, row_data = NULL) {
    # Validate inputs
    if (!is.matrix(counts)) {
        counts <- as.matrix(counts)
    }
    
    if (!is.data.frame(col_data)) {
        stop("col_data must be a data.frame")
    }
    
    # Check that sample IDs match
    if (!all(colnames(counts) %in% rownames(col_data))) {
        stop("Column names of counts must match row names of col_data")
    }
    
    # Reorder col_data to match counts column order
    col_data <- col_data[colnames(counts), , drop = FALSE]
    
    # Build the SummarizedExperiment
    if (is.null(row_data)) {
        se <- SummarizedExperiment(
            assays = list(counts = counts),
            colData = col_data
        )
    } else {
        row_data <- row_data[rownames(counts), , drop = FALSE]
        se <- SummarizedExperiment(
            assays = list(counts = counts),
            colData = col_data,
            rowData = row_data
        )
    }
    
    return(se)
}
```

### Testing `make_se()`

Let's verify our function works by recreating a SummarizedExperiment from the airway data:

```{r test-make-se}
# Extract components from airway
test_counts <- assay(airway, "counts")
test_meta <- as.data.frame(colData(airway))

# Create a new SE using our function
my_se <- make_se(test_counts, test_meta)

# Verify it worked
my_se
dim(my_se)
colData(my_se)
```

> **Exercise B:** Modify the call to `make_se()` to also include `rowData`. Extract it from `airway` first, then pass it as the `row_data` argument.

---

## Function 2: `top_variable_features()` — Select Most Variable Genes

For PCA, we typically want to focus on the most variable genes:

```{r top-variable}
#' Select top variable features
#'
#' @param se A SummarizedExperiment object
#' @param n Number of top variable features to select (default: 500)
#' @param assay_name Name of assay to use (default: "counts")
#'
#' @return A SummarizedExperiment subset to the top n variable features
top_variable_features <- function(se, n = 500, assay_name = "counts") {
    # Get the assay data
    mat <- assay(se, assay_name)
    
    # Calculate variance for each gene (row)
    vars <- apply(mat, 1, var)
    
    # Get indices of top n most variable
    top_idx <- order(vars, decreasing = TRUE)[seq_len(min(n, length(vars)))]
    
    # Subset the SummarizedExperiment
    se[top_idx, ]
}
```

```{r test-top-var}
# Get top 500 variable genes
se_top <- top_variable_features(airway, n = 500)
dim(se_top)
```

> **Discussion:** Why do we return the subsetted `SummarizedExperiment` rather than just a matrix of values? What advantage does this provide?

---

## Function 3: `run_pca()` — Perform PCA

Now let's run PCA on our data:

```{r run-pca}
#' Run PCA on a SummarizedExperiment
#'
#' @param se A SummarizedExperiment object
#' @param assay_name Name of assay to use (default: "counts")
#' @param n_top Number of top variable features to use (default: 500)
#' @param scale Logical; should features be scaled? (default: TRUE)
#' @param log_transform Logical; should counts be log-transformed? (default: TRUE)
#'
#' @return A list with:
#'   - pca: The prcomp object
#'   - scores: A data.frame of PC scores merged with sample metadata
run_pca <- function(se, assay_name = "counts", n_top = 500, 
                    scale = TRUE, log_transform = TRUE) {
    # Subset to top variable features
    se_top <- top_variable_features(se, n = n_top, assay_name = assay_name)
    
    # Get the data matrix
    mat <- assay(se_top, assay_name)
    
    # Log-transform if requested (add pseudocount to avoid log(0))
    if (log_transform) {
        mat <- log2(mat + 1)
    }
    
    # Transpose: prcomp expects samples as rows
    mat_t <- t(mat)
    
    # Run PCA
    pca_result <- prcomp(mat_t, scale. = scale, center = TRUE)
    
    # Create scores data.frame with sample metadata
    scores <- as.data.frame(pca_result$x)
    scores$sample_id <- rownames(scores)
    
    # Merge with colData
    col_data <- as.data.frame(colData(se))
    col_data$sample_id <- rownames(col_data)
    scores <- merge(scores, col_data, by = "sample_id")
    
    return(list(
        pca = pca_result,
        scores = scores
    ))
}
```

```{r test-pca}
# Run PCA on airway data
pca_result <- run_pca(airway, n_top = 500)

# Examine the scores
head(pca_result$scores)

# Check variance explained
summary(pca_result$pca)
```

---

## Function 4: `pca_variance_explained()` — Helper for Variance

```{r pca-variance}
#' Get variance explained by each PC
#'
#' @param pca_result Output from run_pca()
#'
#' @return A data.frame with PC names and percent variance explained
pca_variance_explained <- function(pca_result) {
    pca <- pca_result$pca
    var_explained <- pca$sdev^2 / sum(pca$sdev^2) * 100
    
    data.frame(
        PC = paste0("PC", seq_along(var_explained)),
        variance_percent = var_explained
    )
}
```

```{r test-variance}
var_df <- pca_variance_explained(pca_result)
head(var_df)
```

---

## Function 5: `plot_pca()` — Visualize PCA Results

```{r plot-pca}
#' Create a PCA scatter plot
#'
#' @param pca_result Output from run_pca()
#' @param color_by Column name from colData to color points by
#' @param shape_by Optional column name to map to point shape
#' @param pcs Which PCs to plot (default: c(1, 2))
#' @param point_size Size of points (default: 4)
#'
#' @return A ggplot object
plot_pca <- function(pca_result, color_by = NULL, shape_by = NULL, 
                     pcs = c(1, 2), point_size = 4) {
    scores <- pca_result$scores
    var_exp <- pca_variance_explained(pca_result)
    
    # Build PC column names
    pc_x <- paste0("PC", pcs[1])
    pc_y <- paste0("PC", pcs[2])
    
    # Get variance percentages for axis labels
    var_x <- round(var_exp$variance_percent[pcs[1]], 1)
    var_y <- round(var_exp$variance_percent[pcs[2]], 1)
    
    # Build the plot
    p <- ggplot(scores, aes(x = .data[[pc_x]], y = .data[[pc_y]])) +
        theme_minimal(base_size = 14) +
        labs(
            x = paste0(pc_x, " (", var_x, "% variance)"),
            y = paste0(pc_y, " (", var_y, "% variance)"),
            title = "PCA Plot"
        )
    
    # Add color aesthetic if specified
    if (!is.null(color_by)) {
        p <- p + aes(color = .data[[color_by]])
    }
    
    # Add shape aesthetic if specified
    if (!is.null(shape_by)) {
        p <- p + aes(shape = .data[[shape_by]])
    }
    
    # Add points
    p <- p + geom_point(size = point_size)
    
    return(p)
}
```

### Creating PCA Visualizations

```{r viz-pca}
# Basic PCA plot colored by treatment
plot_pca(pca_result, color_by = "dex")
```

```{r viz-pca-multi}
# Color by treatment, shape by cell line
plot_pca(pca_result, color_by = "dex", shape_by = "cell")
```

> **Exercise C:** Create a PCA plot showing PC2 vs PC3 instead of PC1 vs PC2. What do you observe?

---

# Visualization with ComplexHeatmap

Let's visualize the top variable genes using a heatmap. This is a common way to explore patterns in gene expression data.

```{r heatmap-prep}
# Get top 20 most variable genes for visualization
se_top20 <- top_variable_features(airway, n = 20)

# Extract and log-transform the counts
heatmap_mat <- log2(assay(se_top20, "counts") + 1)

# Scale by row (gene) for better visualization
heatmap_mat <- t(scale(t(heatmap_mat)))
```

```{r heatmap-basic}
# Create annotation for columns (samples)
col_anno <- HeatmapAnnotation(
    treatment = airway$dex,
    cell_line = airway$cell,
    col = list(
        treatment = c("trt" = "steelblue", "untrt" = "salmon"),
        cell_line = c("N052611" = "#E41A1C", "N061011" = "#377EB8", 
                      "N080611" = "#4DAF4A", "N61311" = "#984EA3")
    )
)

# Create the heatmap
Heatmap(
    heatmap_mat,
    name = "Z-score",
    top_annotation = col_anno,
    show_row_names = TRUE,
    show_column_names = TRUE,
    column_title = "Top 20 Variable Genes",
    row_names_gp = gpar(fontsize = 8)
)
```

> **Exercise D:** Modify the code to show the top 50 variable genes. You may need to adjust `show_row_names` for readability.

---

# Summary: The Analysis Core

We now have five core functions that work together:

| Function | Purpose | Input | Output |
|----------|---------|-------|--------|
| `make_se()` | Create SE from raw data | counts matrix + metadata | SummarizedExperiment |
| `top_variable_features()` | Select variable genes | SE + n | Subsetted SE |
| `run_pca()` | Perform PCA | SE + parameters | list(pca, scores) |
| `pca_variance_explained()` | Get variance stats | PCA result | data.frame |
| `plot_pca()` | Visualize PCA | PCA result + aesthetics | ggplot |

These functions:

- **Take structured objects as input** — Not loose matrices that can get out of sync
- **Return well-defined outputs** — Predictable return types make them composable
- **Are small and testable** — Each does one thing well
- **Are reusable** — They don't depend on global variables or specific file paths

---

# After-Class Tasks

## Micro-task 1: Create `analysis_core.R`

Put all five functions into a single script called `analysis_core.R` with:

- Clear function signatures
- Short comments describing inputs and outputs
- No `library()` calls inside the functions (we'll handle dependencies properly in Lecture 5)

## Micro-task 2: Reflection

Write 3-5 sentences describing why a structured container (`SummarizedExperiment`) is preferable to passing 3 separate `data.frames` around. Consider:

- What happens when you subset?
- What happens when you reorder?
- How does it affect function signatures?

## Reading

- Advanced R sections on OOP + S4 (alignment workbook links)
- SummarizedExperiment vignette (run `vignette("SummarizedExperiment")`)

---

# Appendix: Complete Function Reference

```{r all-functions, eval=FALSE}
# SAVE THIS AS: analysis_core.R

#' Create a SummarizedExperiment from counts and metadata
make_se <- function(counts, col_data, row_data = NULL) {
    if (!is.matrix(counts)) counts <- as.matrix(counts)
    if (!is.data.frame(col_data)) stop("col_data must be a data.frame")
    if (!all(colnames(counts) %in% rownames(col_data))) {
        stop("Column names of counts must match row names of col_data")
    }
    col_data <- col_data[colnames(counts), , drop = FALSE]
    if (is.null(row_data)) {
        SummarizedExperiment(assays = list(counts = counts), colData = col_data)
    } else {
        row_data <- row_data[rownames(counts), , drop = FALSE]
        SummarizedExperiment(
            assays = list(counts = counts), 
            colData = col_data, 
            rowData = row_data
        )
    }
}

#' Select top variable features
top_variable_features <- function(se, n = 500, assay_name = "counts") {
    mat <- assay(se, assay_name)
    vars <- apply(mat, 1, var)
    top_idx <- order(vars, decreasing = TRUE)[seq_len(min(n, length(vars)))]
    se[top_idx, ]
}

#' Run PCA on a SummarizedExperiment
run_pca <- function(se, assay_name = "counts", n_top = 500, 
                    scale = TRUE, log_transform = TRUE) {
    se_top <- top_variable_features(se, n = n_top, assay_name = assay_name)
    mat <- assay(se_top, assay_name)
    if (log_transform) mat <- log2(mat + 1)
    mat_t <- t(mat)
    pca_result <- prcomp(mat_t, scale. = scale, center = TRUE)
    scores <- as.data.frame(pca_result$x)
    scores$sample_id <- rownames(scores)
    col_data <- as.data.frame(colData(se))
    col_data$sample_id <- rownames(col_data)
    scores <- merge(scores, col_data, by = "sample_id")
    list(pca = pca_result, scores = scores)
}

#' Get variance explained by each PC
pca_variance_explained <- function(pca_result) {
    pca <- pca_result$pca
    var_explained <- pca$sdev^2 / sum(pca$sdev^2) * 100
    data.frame(PC = paste0("PC", seq_along(var_explained)), 
               variance_percent = var_explained)
}

#' Create a PCA scatter plot
plot_pca <- function(pca_result, color_by = NULL, shape_by = NULL, 
                     pcs = c(1, 2), point_size = 4) {
    scores <- pca_result$scores
    var_exp <- pca_variance_explained(pca_result)
    pc_x <- paste0("PC", pcs[1])
    pc_y <- paste0("PC", pcs[2])
    var_x <- round(var_exp$variance_percent[pcs[1]], 1)
    var_y <- round(var_exp$variance_percent[pcs[2]], 1)
    p <- ggplot(scores, aes(x = .data[[pc_x]], y = .data[[pc_y]])) +
        theme_minimal(base_size = 14) +
        labs(x = paste0(pc_x, " (", var_x, "% variance)"),
             y = paste0(pc_y, " (", var_y, "% variance)"),
             title = "PCA Plot")
    if (!is.null(color_by)) p <- p + aes(color = .data[[color_by]])
    if (!is.null(shape_by)) p <- p + aes(shape = .data[[shape_by]])
    p + geom_point(size = point_size)
}
```

---

# Session Info

```{r session-info}
sessionInfo()
```
