---
title: "Lecture 11: Review / Q&A (Homework 1 Support)"
author: "Jared Andrews"
date: "April 20, 2026"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  eval = FALSE
)
```

# Learning Objectives

By the end of this session, you will be able to:

1. Identify and resolve remaining blockers for Homework 1 submission
2. Apply a systematic debugging workflow (reproduce → isolate → fix → test → document)
3. Validate the end-to-end deliverable: install from GitHub, run core functions, run Shiny app, run CLI, and view pkgdown site

**Course Learning Outcomes (CLOs):** CLO 1, 2, 3, 4, 5, 6

---

# Homework 1 Requirements Recap

## What You're Submitting

A GitHub repository containing an R package that:

| Component | Status | Check |
|-----------|--------|-------|
| **Package structure** | Valid R package | `devtools::check()` passes |
| **Core functions** | `run_pca()`, `plot_pca()`, etc. | Functions work on test data |
| **Documentation** | roxygen2 + README | `?function` works |
| **Tests** | testthat | `devtools::test()` passes |
| **pkgdown site** | Builds successfully | `pkgdown::build_site()` works |
| **Shiny app** | `run_app()` | App launches and works |
| **CLI** | `sePCA` (Rapp CLI) | CLI runs from fresh install |
| **GitHub** | Public repo | `install_github()` works |

---

# Part 1: Readiness Checklist (10 min)

## Run the Checklist

Everyone run through this checklist right now. Mark each item ✅ or ❌:

### 1. Install from GitHub

```{r checklist-1}
# Start a fresh R session (important!)
.rs.restartR()

# Remove local version if exists
try(remove.packages("sePCA"), silent = TRUE)

# Install from GitHub
remotes::install_github("YOUR-USERNAME/sePCA")

# ✅ Success / ❌ Failed
```

### 2. R CMD check

```{r checklist-2}
devtools::check()

# Goal: 0 errors, 0 warnings
# NOTEs are usually acceptable
# ✅ 0 errors, 0 warnings / ❌ Has errors or warnings
```

### 3. Documentation

```{r checklist-3}
library(sePCA)

# Help files work?
?run_pca
?plot_pca

# Package help?
?sePCA

# ✅ Help displays / ❌ Missing docs
```

### 4. Tests

```{r checklist-4}
devtools::test()

# Goal: All tests pass
# ✅ All passed / ❌ Failures
```

### 5. pkgdown Site

```{r checklist-5}
pkgdown::build_site()

# Opens in browser - check:
# - Home page renders
# - Reference pages work
# - Articles/vignettes render

# ✅ Site builds / ❌ Build fails
```

### 6. Shiny App

```{r checklist-6}
library(sePCA)
run_app()

# Check:
# - App launches without error
# - Plot renders
# - Inputs change the plot

# ✅ App works / ❌ App fails
```

### 7. CLI

```bash
# From terminal
sePCA pca --help

# If launchers are not installed
# Rapp exec/sePCA pca --help

# Should show help text
# ✅ Help shows / ❌ Error
```

---

# Part 2: Triage (10 min)

## Categorize Your Blockers

Based on your checklist results, identify which category your issues fall into:

| Category | Examples |
|----------|----------|
| **Package/Check** | R CMD check errors, missing NAMESPACE entries, dependency issues |
| **Documentation** | Missing roxygen2 tags, broken examples, README issues |
| **Tests/CI** | Failing tests, CI not running, test setup issues |
| **pkgdown** | Site build errors, missing vignettes, config issues |
| **Shiny** | App crashes, missing dependencies, broken reactivity |
| **CLI** | Argument parsing, file I/O, exit codes |

## Form Groups

We'll form debugging groups by category. Join the group matching your primary blocker.

---

# Part 3: Debugging Playbook

## The Systematic Approach

```
┌─────────────┐     ┌─────────────┐     ┌─────────────┐
│  Reproduce  │ ──→ │   Isolate   │ ──→ │     Fix     │
└─────────────┘     └─────────────┘     └─────────────┘
        │                                       │
        └───────────────────────────────────────┘
                         ↓
                 ┌─────────────┐     ┌─────────────┐
                 │    Test     │ ──→ │  Document   │
                 └─────────────┘     └─────────────┘
```

### 1. Reproduce

Can you reliably trigger the error?

```{r reproduce}
# Write down exact steps:
# 1. Restart R
# 2. Run: library(sePCA)
# 3. Run: run_pca(example_se)
# 4. Error appears: "could not find function..."
```

### 2. Isolate

What's the minimal code that triggers it?

```{r isolate}
# Start removing parts until you find the core issue
# Does this work?
library(SummarizedExperiment)
SummarizedExperiment(...)

# Does this fail?
sePCA::run_pca(...)
```

### 3. Fix

Apply the fix:

```{r fix}
# Common fixes by error type

# "could not find function X" 
# → Add @importFrom or use pkg::fun()

# "missing argument" 
# → Add @param tags in roxygen

# "no visible binding for global variable"
# → Use .data$variable in ggplot aes()
```

### 4. Test

Verify the fix:

```{r test}
# Run the minimal reproduction case
# Confirm error is gone
# Run full test suite
devtools::test()
```

### 5. Document

Record what you learned:

```{r document}
# Add a comment explaining the fix
# Update documentation if needed
# Commit with descriptive message:
# "Fix: Add SummarizedExperiment to Imports for run_pca()"
```

---

# Common Issues and Fixes

## Package/Check Errors

### "could not find function X"

**Problem:** Function from another package isn't available.

**Fix:**

```{r fix-missing-function}
# Option 1: Explicit namespace
SummarizedExperiment::assay(se, "counts")

# Option 2: Import via roxygen2
#' @importFrom SummarizedExperiment assay
```

### "no visible binding for global variable"

**Problem:** ggplot2 aesthetics trigger R CMD check notes.

**Fix:**

```{r fix-binding}
# Instead of:
ggplot(data, aes(x = PC1, y = PC2))

# Use:
ggplot(data, aes(x = .data$PC1, y = .data$PC2))

# Or add to package (in R/globals.R):
utils::globalVariables(c("PC1", "PC2"))
```

### Missing imports in DESCRIPTION

**Fix:**

```{r fix-imports}
# Add missing packages
usethis::use_package("ggplot2")
usethis::use_package("SummarizedExperiment")
```

---

## Documentation Issues

### "Undocumented arguments"

**Fix:** Add missing `@param` tags:

```{r fix-params}
#' @param se A SummarizedExperiment object
#' @param n_top Number of top variable genes
```

### Examples fail

**Fix:** Wrap long-running or interactive examples:

```{r fix-examples}
#' @examples
#' \dontrun{
#' # This example takes a while
#' run_pca(large_dataset)
#' }
#'
#' \donttest{
#' # This is tested but not on CRAN
#' run_app()
#' }
```

---

## Test Issues

### Test can't find package functions

**Fix:** Make sure you're using the package correctly in tests:

```{r fix-test-loading}
# tests/testthat/test-pca.R

# Functions should be available without library()
# But you may need to load package data:
test_that("run_pca works", {
    data(example_se, package = "sePCA")  # Explicit package
    result <- run_pca(example_se)
    expect_s3_class(result$scores, "data.frame")
})
```

### Tests fail on CI but pass locally

**Possible causes:**

1. Missing Suggests dependency
2. Different R version
3. Missing system dependency
4. Hardcoded paths

**Fix:** Check CI logs and ensure all test dependencies are in Suggests.

---

## pkgdown Issues

### Site won't build

**Common fixes:**

```{r fix-pkgdown}
# Check pkgdown config is valid YAML
yaml::read_yaml("_pkgdown.yml")

# Rebuild from scratch
pkgdown::clean_site()
pkgdown::build_site()

# Check for vignette issues
devtools::build_vignettes()
```

### Missing reference pages

**Fix:** Ensure functions are exported:

```{r fix-export}
#' @export
run_pca <- function(...) {
  ...
}
```

---

## Shiny Issues

### App crashes on startup

**Debug approach:**

```{r debug-shiny}
# Run with error tracing
options(shiny.error = browser)
run_app()

# Or check the app components separately
app_ui()      # Does UI build?
# Server can't be tested as easily, but:
devtools::load_all()  # Make sure package loads
```

### Reactivity not working

**Common causes:**

```{r debug-reactivity}
# Forgot parentheses on reactive
output$plot <- renderPlot({
    plot_pca(pca_result)  # WRONG: missing ()
    plot_pca(pca_result())  # RIGHT
})

# Wrong input name
input$n_top  # Check this matches UI: numericInput("n_top", ...)
```

---

## CLI Issues

### Arguments not parsing

**Debug:**

```{r debug-cli}
# Test the CLI directly from R
Rapp::run(
    system.file("exec", "sePCA", package = "sePCA"),
    c("pca", "--counts", "test.tsv", "--meta", "meta.tsv", "--output", "out/")
)
```

### Files not found

**Check:**

```{r debug-files}
# Are you using absolute paths?
# Relative paths are relative to working directory!
getwd()

# Better: require absolute paths
normalizePath(args$counts)
```

---

# Part 4: Show and Tell (20 min)

## Share Your Wins

2-3 volunteers will demo:

1. What problem they had
2. How they diagnosed it
3. What the fix was

## Capture Patterns

As we discuss, we'll build a shared FAQ:

```markdown
## FAQ: Common Issues and Fixes

### Q: My function works locally but not after install
A: Make sure you're not using `devtools::load_all()` logic. Test with 
a fresh `remotes::install_github()` install.

### Q: R CMD check complains about global variables
A: Use `.data$varname` in ggplot2 aes() calls, or add `globalVariables()`.

### Q: CI fails but local passes
A: Check that all dependencies are in DESCRIPTION. Look at the CI log
for the specific error.

... (add more as we discuss)
```

---

# Part 5: Peer Review Exercise

## Exercise A: Minimal Bug Report

For one unresolved blocker, write an issue-style bug report:

```markdown
## Bug: [Short description]

### Expected behavior
[What should happen]

### Actual behavior
[What actually happens]

### Steps to reproduce
1. Start a fresh R session
2. Run: `library(sePCA)`
3. Run: `run_pca(example_se)`
4. See error: [error message]

### Environment
```r
sessionInfo()
# Paste output here
```

### Additional context
[Screenshots, log files, etc.]
```

## Exercise B: Swap and Review

1. Pair up with a classmate
2. Give them your GitHub repo URL
3. They attempt to:
   - Install from GitHub
   - Run the README example
   - Run `run_app()`
   - Run the CLI

4. They file one constructive GitHub issue:
   - Be specific about what doesn't work
   - Include exact error messages
   - Suggest a fix if you can

---

# Final Submission Checklist

## Before You Submit

```markdown
## HW1 Submission Checklist

### Package Basics
- [ ] Package installs from GitHub without errors
- [ ] `devtools::check()` passes with 0 errors, 0 warnings
- [ ] All exported functions have roxygen2 documentation
- [ ] README exists and renders on GitHub

### Core Functionality
- [ ] `run_pca()` returns list with `pca` and `scores`
- [ ] `plot_pca()` returns a ggplot object

### Testing
- [ ] At least 5 test expectations across 2+ files
- [ ] All tests pass with `devtools::test()`
- [ ] CI runs on GitHub (Actions badge shows)

### Shiny App
- [ ] `run_app()` launches without errors
- [ ] Plot updates when inputs change
- [ ] App can be run from fresh R session

### CLI
- [ ] `exec/sePCA` exists in the package
- [ ] `Rapp::install_pkg_cli_apps("sePCA")` installs a launcher
- [ ] `--help` shows usage information
- [ ] PCA command produces output files

### Documentation
- [ ] pkgdown site builds successfully
- [ ] At least one vignette/article exists
- [ ] README has installation and quick start

### Final Verification
- [ ] From fresh R: `remotes::install_github("user/repo")` works
- [ ] From fresh R: `library(pkg); run_app()` works
- [ ] From terminal: CLI produces expected outputs
```

## Optional Enhancements

- [ ] Tag a release (v0.1.0)
- [ ] Deploy Shiny app to shinyapps.io
- [ ] Deploy pkgdown site to GitHub Pages
- [ ] Add CI badges to README

---

# Summary

Today we:

1. Ran a comprehensive readiness checklist
2. Triaged issues by category
3. Applied systematic debugging workflows
4. Shared solutions and built a FAQ
5. Practiced peer code review
6. Finalized submission checklists

## You're Ready!

If you've completed the checklist, your package is ready for submission.

---

# Resources

## Getting Help

- **Stack Overflow**: Tag with `r`, `r-package-devel`
- **RStudio Community**: https://community.rstudio.com/
- **R-pkg-devel mailing list**: For package development questions
- **Course Slack/Discord**: Ask classmates!

## Essential References

- **R Packages book**: https://r-pkgs.org/
- **Mastering Shiny**: https://mastering-shiny.org/
- **pkgdown docs**: https://pkgdown.r-lib.org/
- **testthat docs**: https://testthat.r-lib.org/

## Example Packages to Study

- **usethis**: Great example of devtools conventions
- **ggplot2**: Excellent documentation
- **shiny**: See how they structure the app

---

# Session Info

```{r session-info, eval=TRUE}
sessionInfo()
```
