---
title: "Lecture 5: Lab – R Package Development (devtools)"
author: "Jared Andrews"
date: "April 15, 2026"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  eval = FALSE  # Don't actually run package creation code when knitting
)
```

# Learning Objectives

By the end of this session, you will be able to:

1. Create an R package skeleton with usethis and organize code into the correct package structure
2. Define and validate a correct DESCRIPTION file, including dependencies and metadata
3. Add roxygen2 documentation and generate help files/NAMESPACE
4. Publish the package to a Git repository and install it from GitHub
5. Diagnose and resolve common development issues (namespacing, missing Imports, R CMD check warnings)

**Course Learning Outcomes (CLOs):** CLO 1, 4, 6

---

# Why Package Your Code?

## From Script to Application

In Lecture 3, we created an "analysis core" — a set of functions in a script. But scripts have limitations:

| Script | Package |
|--------|---------|
| `source("path/to/file.R")` | `library(sePCA)` |
| Paths break when you move files | Installed; works anywhere |
| Dependency conflicts | Explicit dependency management |
| No help documentation | `?run_pca` works |
| Hard to share | `install_github("you/sePCA")` |
| No tests | Automated testing with testthat |

**A package is the fundamental unit of shareable, reproducible code in R.**

---

# Pre-Lab Checklist

Before we begin, ensure you have:

- [ ] Your `analysis_core.R` script from Lecture 3 with the five functions
- [ ] Git installed and configured (`git config --global user.name "Your Name"`)
- [ ] A GitHub account
- [ ] RStudio (recommended) or another IDE with R support

```{r install-dev-tools, eval=FALSE}
# Install development tools if needed
install.packages(c("devtools", "usethis", "roxygen2", "testthat"))
```

---

# Part 1: Define the Package API

Before writing code, let's define the **user story** and **public API**.

## User Story

> As a bioinformatics researcher, I want to quickly run PCA on a SummarizedExperiment and create publication-ready visualizations, so that I can explore batch effects and sample groupings in my RNA-seq data.

## Public API (Exported Functions)

| Function | Exported? | Reason |
|----------|-----------|--------|
| `top_variable_features()` | Yes | Users may want to filter separately |
| `run_pca()` | Yes | Core analysis function |
| `pca_variance_explained()` | Yes | Useful standalone |
| `plot_pca()` | Yes | Core visualization |

> **Exercise A:** Are there any functions you might add later that should be **internal** (not exported)? What distinguishes an exported function from an internal helper?

---

# Part 2: Create the Package Skeleton

## Step 1: Create the Package

Open RStudio and create a new package:

```{r create-package}
library(usethis)

# Create the package (choose your own name!)
# This creates a new directory with the package structure
create_package("~/sePCA")  # Or wherever you want it

# This will open a new RStudio session in the package directory
```

After running this, you'll see:

```
sePCA/
├── .Rbuildignore
├── .gitignore
├── DESCRIPTION
├── NAMESPACE
├── R/
└── sePCA.Rproj
```

## Step 2: Initialize Git

```{r init-git}
# Initialize git repository
use_git()

# This will:
# 1. Create a .git directory
# 2. Make an initial commit
# 3. Potentially restart RStudio
```

## Step 3: Connect to GitHub

```{r use-github}
# Create a GitHub repository and push
use_github()

# This will:
# 1. Create a new repo on GitHub
# 2. Add the remote
# 3. Push your initial commit

# If this doesn't work, you can create the repo manually on GitHub
# and then add the remote via the terminal:
# git remote add origin https://github.com/your-username/sePCA.git
# git push -u origin main
```

## Step 4: Add a License

```{r add-license}
# MIT License is a good default for open source
use_mit_license()

# This adds:
# - LICENSE.md file
# - License field in DESCRIPTION
```

---

# Part 3: The DESCRIPTION File

The `DESCRIPTION` file is the heart of your package metadata. Open it and edit:

```yaml
Package: sePCA
Title: PCA Analysis for SummarizedExperiment Objects
Version: 0.0.0.9000
Authors@R: 
    person("Your", "Name", email = "you@example.com", role = c("aut", "cre"),
           comment = c(ORCID = "YOUR-ORCID-ID"))
Description: Provides functions to perform principal component analysis (PCA)
    on SummarizedExperiment objects. Includes tools for selecting top variable
    features, running PCA, and creating publication-ready visualizations.
    Designed as a teaching example for scientific application development.
License: MIT + file LICENSE
Encoding: UTF-8
Roxygen: list(markdown = TRUE)
RoxygenNote: 7.2.3
```

## Adding Dependencies

**Critical:** In packages, we don't use `library()`. Instead, we declare dependencies in DESCRIPTION.

```{r add-dependencies}
# Add packages to Imports (required to run)
use_package("SummarizedExperiment")
use_package("ggplot2")
use_package("rlang")  # For .data pronoun in ggplot aes()

# Add packages to Suggests (optional, for examples/tests)
use_package("testthat", type = "Suggests")
use_package("knitr", type = "Suggests")

# NOTE: For Bioconductor packages, you may also need:
# BiocManager::install("SummarizedExperiment")
# and add 'biocViews:' to your DESCRIPTION (can be left empty)
```

After running these, your DESCRIPTION will include:

```yaml
Imports:
    ggplot2,
    SummarizedExperiment
Suggests:
    knitr,
    testthat
```

> **Exercise B:** Why do we put `testthat` in Suggests rather than Imports? What's the difference?

---

# Part 4: Move Functions to R/

## Package Structure for Code

All R code goes in the `R/` directory. You can organize it however you like, but a common pattern:

```
R/
├── data.R         # Functions for data handling (top_variable_features)
├── pca.R          # PCA-related functions
├── plotting.R     # Visualization functions
└── sePCA-package.R  # Package-level documentation
```

## Create the Files

```{r create-r-files}
# Create R files (this just opens them in RStudio)
use_r("data")      # For top_variable_features()
use_r("pca")       # For run_pca(), pca_variance_explained()
use_r("plotting")  # For plot_pca()
```

## Move and Adapt Your Functions

Copy your functions from `analysis_core.R`, but make these changes:

### Before (Script Style)

```{r script-style, eval=FALSE}
library(SummarizedExperiment)  # DON'T do this in packages!
library(ggplot2)

top_variable_features <- function(se, n = 500, assay_name = "counts") {
  mat <- assay(se, assay_name)  # Uses assay() directly
  # ...
}
```

### After (Package Style)

Two options for using functions from other packages:

**Option 1: Full namespace (recommended for clarity)**

```{r namespace-style, eval=FALSE}
# In R/data.R
top_variable_features <- function(se, n = 500, assay_name = "counts") {
  # Use pkg::fun() syntax
  mat <- SummarizedExperiment::assay(se, assay_name)
  # ...
}
```

**Option 2: Import via roxygen2**

```{r import-style, eval=FALSE}
#' @importFrom SummarizedExperiment assay colData rowData
top_variable_features <- function(se, n = 500, assay_name = "counts") {
  # Can use function directly because it's imported
  mat <- assay(se, assay_name)
  # ...
}
```

### Complete Example: R/data.R

```{r data-r}
# R/data.R

#' Select top variable features
#'
#' @param se A SummarizedExperiment object
#' @param n Number of top variable features to select (default: 500)
#' @param assay_name Name of assay to use (default: "counts")
#'
#' @return A SummarizedExperiment subset to the top n variable features
#' @export
#'
#' @examples
#' # Assuming 'se' is a SummarizedExperiment
#' # se_top <- top_variable_features(se, n = 500)
top_variable_features <- function(se, n = 500, assay_name = "counts") {
    mat <- SummarizedExperiment::assay(se, assay_name)
    vars <- apply(mat, 1, var)
    top_idx <- order(vars, decreasing = TRUE)[seq_len(min(n, length(vars)))]
    se[top_idx, ]
}
```

Note the key additions:

- `#' @export` — Makes the function available to users
- `#' @param` — Documents each parameter
- `#' @return` — Documents the return value
- `#' @examples` — Provides runnable examples
- `pkg::fun()` syntax — Explicit namespacing

---

# Part 5: Generate Documentation

## Run roxygen2

```{r document}
library(devtools)

# Generate documentation and NAMESPACE
document()
```

This creates:

- `man/top_variable_features.Rd` — Help file for each documented function
- `NAMESPACE` — Updated with exports and imports

Check your `NAMESPACE` file:

```
# Generated by roxygen2: do not edit by hand

export(pca_variance_explained)
export(plot_pca)
export(run_pca)
export(top_variable_features)
```

## Test the Help

```{r test-help}
# Load the package in development mode
load_all()

# Test help pages
?top_variable_features
?run_pca
```

---

# Part 6: Add Example Data

Let's include a small example dataset so users (and tests) can try the functions without loading external data.

```{r data-raw}
# Create data-raw directory for data preparation scripts
use_data_raw("example_se")
```

This creates `data-raw/example_se.R`. Edit it:

```{r example-data-script}
# data-raw/example_se.R

# Create a small example SummarizedExperiment
library(SummarizedExperiment)

set.seed(42)

# 100 genes, 8 samples
n_genes <- 100
n_samples <- 8

# Simulate counts (negative binomial-ish)
counts <- matrix(
    rpois(n_genes * n_samples, lambda = 100),
    nrow = n_genes,
    ncol = n_samples
)
rownames(counts) <- paste0("gene", seq_len(n_genes))
colnames(counts) <- paste0("sample", seq_len(n_samples))

# Add some structure: first 20 genes differ by treatment
treatment <- rep(c("control", "treated"), each = 4)
counts[1:20, treatment == "treated"] <- counts[1:20, treatment == "treated"] * 2

# Sample metadata
sample_data <- data.frame(
    sample_id = colnames(counts),
    treatment = treatment,
    batch = rep(c("A", "B"), times = 4),
    row.names = colnames(counts)
)

# Gene metadata
gene_data <- data.frame(
    gene_id = rownames(counts),
    gene_symbol = paste0("SYM", seq_len(n_genes)),
    row.names = rownames(counts)
)

# Create SummarizedExperiment
example_se <- SummarizedExperiment(
    assays = list(counts = counts),
    colData = sample_data,
    rowData = gene_data
)

# Save to data/
usethis::use_data(example_se, overwrite = TRUE)
```

Run the script to create the data:

```{r run-data-script}
source("data-raw/example_se.R")
```

Now document the data:

```{r document-data}
use_r("data-documentation")
```

Add to `R/data-documentation.R`:

```{r data-docs}
#' Example SummarizedExperiment for testing
#'
#' A small SummarizedExperiment with 100 genes and 8 samples.
#' Includes a treatment effect in the first 20 genes.
#'
#' @format A SummarizedExperiment with:
#' \describe{
#'   \item{assays}{counts - raw count matrix}
#'   \item{colData}{sample_id, treatment (control/treated), batch (A/B)}
#'   \item{rowData}{gene_id, gene_symbol}
#' }
#'
#' @source Simulated data for teaching purposes
#'
#' @examples
#' data(example_se)
#' example_se
#' colData(example_se)
"example_se"
```

---

# Part 7: R CMD check

This is the moment of truth! `R CMD check` runs ~50 checks to validate your package.

```{r check}
# Run the full check
check()
```

## Common Issues and Fixes

### 1. Missing Imports

```
Error: could not find function "SummarizedExperiment"
```

**Fix:** Add `#' @importFrom` or use `pkg::fun()` syntax.

### 2. Undocumented Arguments

```
Warning: Undocumented arguments in documentation object 'top_variable_features'
  'row_data'
```

**Fix:** Add `@param row_data` to your roxygen2 block.

### 3. Non-ASCII Characters

```
Warning: found non-ASCII string
```

**Fix:** Replace special characters (curly quotes, em-dashes) with ASCII equivalents.

### 4. Missing Suggests for Examples

```
Package suggested but not available: 'foo'
```

**Fix:** Wrap example code in `\dontrun{}` or add the package to Suggests.

> **Exercise C:** Run `check()` on your package. Categorize each WARNING and NOTE as either "must fix" or "acceptable for now."

---

# Part 8: Push to GitHub and Test Install

## Commit and Push

```{r git-commit}
# Stage all files
# git add -A

# Commit
# git commit -m "Initial package structure with core PCA functions"

# Push
# git push
```

In RStudio, use the Git pane (Ctrl+Alt+M to commit).

## Test Installation

Open a **fresh R session** (important!) and try:
                    
```{r test-install, eval=FALSE}
# Install from GitHub
# remotes::install_github("your-username/sePCA")

# Load and test
library(sePCA)
data(example_se)

result <- run_pca(example_se)
plot_pca(result, color_by = "treatment")
```

---

# Part 9: Add a README

```{r readme}
use_readme_rmd()
```

Edit `README.Rmd`:

````markdown
---
output: github_document
---

# sePCA

<!-- badges: start -->
<!-- badges: end -->

sePCA provides tools for performing PCA on SummarizedExperiment objects.

## Installation
 
```r
# Install from GitHub
remotes::install_github("your-username/sePCA")
```

## Quick Start

```r
library(sePCA)
library(airway)

# Load example data
data(airway)

# Run PCA
result <- run_pca(airway, n_top = 500)

# Plot
plot_pca(result, color_by = "dex")
```
````

Then build the README:

```{r build-readme}
build_readme()
```

---

# Summary

Today we:

1. Created a package skeleton with `usethis::create_package()`
2. Set up Git and GitHub for version control
3. Configured DESCRIPTION with proper metadata and dependencies
4. Moved our functions to R/ with roxygen2 documentation
5. Created example data for testing
6. Ran `devtools::check()` and fixed issues
7. Published to GitHub and tested installation

## Package Milestone

✅ A valid GitHub-hosted R package that installs successfully and exposes the analysis core as documented exported functions.

---

# After-Class Tasks

## Reading

- Skim the Bioconductor Contributions book sections on package structure
- R Packages book: https://r-pkgs.org (chapters 1-7)

## Micro-task 1: Package-Level Documentation

Add a package-level documentation file:

```{r pkg-docs}
use_package_doc()
```

Edit `R/sePCA-package.R` to describe the package purpose and main functions.

## Micro-task 2: Fresh Install Test

From a fresh R session (restart R first!):

1. Install your package: `remotes::install_github("you/sePCA")`
2. Run the README example
3. Fix any issues

---

# Quick Reference

```{r quick-ref, eval=FALSE}
# Package development workflow
library(devtools)
library(usethis)

# Create package
create_package("path/to/pkgname")

# Add infrastructure
use_git()
use_github()
use_mit_license()
use_readme_rmd()

# Add dependencies
use_package("ggplot2")              # Imports
use_package("testthat", "Suggests") # Suggests

# Create files
use_r("filename")                   # R code
use_data_raw("dataname")            # Data prep script

# Development cycle
load_all()      # Load package for testing (Ctrl+Shift+L)
document()      # Generate docs (Ctrl+Shift+D)
check()         # Run R CMD check (Ctrl+Shift+E)
test()          # Run tests (Ctrl+Shift+T)
build_readme()  # Render README.Rmd

# Install
install()       # Install locally
# install_github("user/repo")  # From GitHub
```

---

# Session Info

```{r session-info, eval=TRUE}
sessionInfo()
```
