---
title: "Lecture 10: Lab – CLI Tool Development (R) – Packaging and Installation"
author: "Jared Andrews"
date: "April 17, 2026"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  eval = FALSE
)
```

# Learning Objectives

By the end of this session, you will be able to:

1. Add the CLI entry point to the package so it works after installation
2. Update the GitHub repository and verify installation and CLI execution in a clean session
3. Confirm that CLI outputs match the package's R function outputs for the same inputs

**Course Learning Outcomes (CLOs):** CLO 1, 2, 3, 5, 6

---

# Where We Are

In Lecture 9, we built CLI functions that work during development:

```bash
# Works in development mode
Rapp exec/sePCA pca --help
```

**Problem:** After installing from GitHub, users need a simpler way to run the CLI.

**Goal:** After `install_github("you/sePCA")`, users can run:

```bash
# Option 1: Launcher (after installing with Rapp)
sePCA pca --counts counts.tsv --meta meta.tsv --output out/

# Option 2: Direct exec (no launcher)
Rapp exec/sePCA pca --counts counts.tsv --meta meta.tsv --output out/
```

---

# Part 1: Package the CLI Entry Point

## Use Rapp's `exec/` Directory

Rapp CLI apps live in a top-level `exec/` directory. Files in `exec/` are installed with the package.

### Create the Directory (if needed)

```{r create-exec}
dir.create("exec", recursive = TRUE)
```

### Add the Rapp App

Create `exec/sePCA` (from Lecture 9). On Unix/Mac, mark it executable:

```bash
chmod +x exec/sePCA
```

### Declare the Dependency

Add Rapp to DESCRIPTION so users have the runner available:

```{r add-rapp}
usethis::use_package("Rapp")
```

---

# Part 2: Make the CLI Discoverable

## Install CLI Launchers

Rapp can install launchers on your PATH:

```r
Rapp::install_pkg_cli_apps("sePCA")
```

After this, users can run:

```bash
sePCA pca --help
```

## Fallback: Run Directly from `exec/`

If launchers are not installed, you can still run the app directly:

```bash
Rapp exec/sePCA pca --help
```

Or from R:

```r
Rapp::run(system.file("exec", "sePCA", package = "sePCA"), c("pca", "--help"))
```

### Optional Helper Function

Add a small helper in `R/cli_install.R` if you want a one-liner for users:

```{r install-helper}
#' Install sePCA CLI launchers
#'
#' @export
install_sePCA_cli <- function(...) {
    Rapp::install_pkg_cli_apps("sePCA", ...)
}
```

---

# Part 3: Documentation

## Update README

Add a CLI section:

````markdown
## Command Line Interface

sePCA includes a command-line interface for pipeline integration.

### Quick Start

```bash
# Install the launcher (one-time)
Rscript -e "Rapp::install_pkg_cli_apps('sePCA')"

# Run PCA analysis
sePCA pca \
    --counts counts.tsv \
    --meta samples.tsv \
    --output results/ \
    --n-top 500 \
    --color-by treatment
```

### Fallback (No Launcher)

```bash
Rapp exec/sePCA pca --help
```

### Available Commands

| Command | Description |
|---------|-------------|
| `pca` | Run PCA analysis and export results |
| `validate` | Validate input files |

Run `sePCA <command> --help` for detailed options.
````

## Add pkgdown Article

Create `vignettes/cli.Rmd`:

````markdown
---
title: "Command Line Interface"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Command Line Interface}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Overview

The sePCA CLI allows you to run PCA analysis from the command line,
making it easy to integrate into pipelines.

## Installation

After installing sePCA, install the CLI launcher:

```r
remotes::install_github("you/sePCA")
Rapp::install_pkg_cli_apps("sePCA")
```

## Usage

### Basic PCA Analysis

```bash
sePCA pca \
    --counts counts.tsv \
    --meta samples.tsv \
    --output results/
```

### Input File Formats

**Counts matrix** (`--counts`):
- Tab-separated or comma-separated
- First column: gene IDs (row names)
- Subsequent columns: sample counts
- Header row with sample IDs

```
gene_id    sample1    sample2    sample3
ENSG001    100        150        120
ENSG002    50         75         60
```

**Metadata** (`--meta`):
- Tab-separated or comma-separated
- First column: sample IDs (must match counts columns)
- Additional columns: sample attributes

```
sample_id    treatment    batch
sample1      control      A
sample2      treated      A
sample3      treated      B
```

### Output Files

| File | Description |
|------|-------------|
| `pca_scores.tsv` | PCA scores merged with sample metadata |
| `pca_variance.tsv` | Variance explained by each PC |
| `pca_plot.png` | PCA scatter plot (if `--color-by` specified) |

### Full Options

```
sePCA pca --help
```

## Pipeline Integration

### Snakemake

```python
rule pca:
    input:
        counts = "data/{sample}/counts.tsv",
        meta = "data/{sample}/meta.tsv"
    output:
        directory("results/{sample}/pca/")
    shell:
        """
        sePCA pca \
            --counts {input.counts} \
            --meta {input.meta} \
            --output {output}
        """
```

### Bash Loop

```bash
for dataset in data/*; do
    name=$(basename "$dataset")
    sePCA pca \
        --counts "$dataset/counts.tsv" \
        --meta "$dataset/meta.tsv" \
        --output "results/$name/"
done
```
````

---

# Part 4: Clean Room Testing

## The Importance of Clean Testing

During development, `devtools::load_all()` makes your code available. But users install the package — we need to verify that works too.

## Exercise A: Clean Room Test

### Step 1: Install from GitHub

In a fresh R session (restart R first!):

```{r clean-install}
# Remove any existing installation
remove.packages("sePCA")

# Install fresh from GitHub
remotes::install_github("you/sePCA")
```

### Step 2: Verify Package Works

```{r verify-pkg}
library(sePCA)

# Test core functions
data(example_se)
result <- run_pca(example_se)
plot_pca(result, color_by = "treatment")

# Test Shiny app launches (close it after)
# run_app()

# Install CLI launcher
Rapp::install_pkg_cli_apps("sePCA")
```

### Step 3: Test CLI from Terminal

```bash
# Create test files
mkdir -p /tmp/cli_test
cat > /tmp/cli_test/counts.tsv << 'EOF'
gene_id	sample1	sample2	sample3	sample4
gene1	100	150	120	180
gene2	50	75	60	90
gene3	200	180	210	240
gene4	30	45	35	55
gene5	80	100	90	110
EOF

cat > /tmp/cli_test/meta.tsv << 'EOF'
sample_id	treatment	batch
sample1	control	A
sample2	control	B
sample3	treated	A
sample4	treated	B
EOF

# Run CLI
sePCA pca \
    --counts /tmp/cli_test/counts.tsv \
    --meta /tmp/cli_test/meta.tsv \
    --output /tmp/cli_test/output/ \
    --n-top 5

# Check outputs
cat /tmp/cli_test/output/pca_scores.tsv
cat /tmp/cli_test/output/pca_variance.tsv
```

---

# Part 5: Output Parity Testing

## What is Output Parity?

The CLI should produce the same results as calling the R functions directly.

## Exercise B: Verify Output Parity

### Step 1: Run Analysis via R

```{r r-analysis}
library(sePCA)

# Load sample data
counts <- read.table("/tmp/cli_test/counts.tsv", header = TRUE, row.names = 1, sep = "\t")
meta <- read.table("/tmp/cli_test/meta.tsv", header = TRUE, row.names = 1, sep = "\t")

# Create SE and run PCA
se <- make_se(as.matrix(counts), meta)
result <- run_pca(se, n_top = 5)

# Get scores
r_scores <- result$scores
```

### Step 2: Compare with CLI Output

```{r compare}
# Load CLI output
cli_scores <- read.table(
    "/tmp/cli_test/output/pca_scores.tsv",
    header = TRUE,
    sep = "\t"
)

# Compare structure
print(names(r_scores))
print(names(cli_scores))

# Compare values (allowing for floating point differences)
all.equal(r_scores$PC1, cli_scores$PC1)
all.equal(r_scores$PC2, cli_scores$PC2)
```

## Adding an Automated Test

Create `tests/testthat/test-cli.R`:

```{r cli-test}
test_that("CLI produces same output as R functions", {
    skip_on_cran()  # CLI tests may be slow
    
    # Create temp files
    tmp_dir <- tempdir()
    counts_file <- file.path(tmp_dir, "counts.tsv")
    meta_file <- file.path(tmp_dir, "meta.tsv")
    output_dir <- file.path(tmp_dir, "cli_output")
    
    # Create test data
    data(example_se, package = "sePCA")
    counts <- SummarizedExperiment::assay(example_se, "counts")
    meta <- as.data.frame(SummarizedExperiment::colData(example_se))
    
    write.table(counts, counts_file, sep = "\t", quote = FALSE)
    write.table(meta, meta_file, sep = "\t", quote = FALSE)
    
    # Run via R
    se <- make_se(counts, meta)
    r_result <- run_pca(se, n_top = 50)
    
    # Run via CLI
    Rapp::run(system.file("exec", "sePCA", package = "sePCA"), c(
        "pca",
        "--counts", counts_file,
        "--meta", meta_file,
        "--output", output_dir,
        "--n-top", "50"
    ))
    
    # Load CLI output
    cli_scores <- read.table(
        file.path(output_dir, "pca_scores.tsv"),
        header = TRUE,
        sep = "\t"
    )
    
    # Compare dimensions
    expect_equal(nrow(cli_scores), nrow(r_result$scores))
    expect_equal(ncol(cli_scores), ncol(r_result$scores))
    
    # Compare PC values (within tolerance)
    expect_equal(cli_scores$PC1, r_result$scores$PC1, tolerance = 1e-10)
    
    # Cleanup
    unlink(output_dir, recursive = TRUE)
})
```

---

# Part 6: Release Checklist

## Pre-Release Checks

Before tagging a release:

```{r pre-release}
# Run all checks
devtools::check()

# Run tests
devtools::test()

# Build documentation
pkgdown::build_site()

# Verify README renders
devtools::build_readme()
```

## Bump Version

```{r bump-version}
# Use usethis to update version
usethis::use_version("minor")  # 0.0.1 -> 0.1.0
```

## Create GitHub Release

```bash
# Tag the release
git tag -a v0.1.0 -m "First release with CLI support"
git push origin v0.1.0
```

On GitHub:
1. Go to Releases → Create a new release
2. Select tag v0.1.0
3. Add release notes:

```markdown
## sePCA v0.1.0

### Features
- Core PCA analysis functions (`run_pca()`, `plot_pca()`)
- Shiny app for interactive exploration (`run_app()`)
- Command-line interface (Rapp app in `exec/`)
- pkgdown documentation site

### Installation
```r
remotes::install_github("you/sePCA@v0.1.0")
```
```

---

## Exercise C: Help Quality

### Peer Review Test

1. Pair up with a classmate
2. Give them only your README and `--help` output
3. Ask them to run the CLI on test data
4. Note any confusion or questions
5. Update documentation to address unclear points

Questions to ask:
- Could they install successfully?
- Was the CLI invocation clear?
- Did they understand what the outputs mean?
- Any error messages that were confusing?

---

# Summary

Today we:

1. Packaged the CLI as a Rapp app in `exec/`
2. Installed launchers with `Rapp::install_pkg_cli_apps()`
3. Updated documentation (README, pkgdown article)
4. Performed clean-room installation testing
5. Verified output parity between R and CLI
6. Created a release checklist

## Package Milestone

OK - The package installs from GitHub and exposes both `run_app()` and a working Rapp-based CLI.

---

# After-Class Tasks

## Micro-task 1: R CMD check

Verify your package passes `devtools::check()` and CI is green.

## Micro-task 2: Reproducibility Section

Add a "Reproducibility" section to README:

```markdown
## Reproducibility

This package was developed with:
- R version 4.3.x
- Bioconductor 3.18

### Installation

```r
# Install from GitHub
remotes::install_github("you/sePCA")

# Or from a specific version
remotes::install_github("you/sePCA@v0.1.0")
```

### Session Info

Run `sessionInfo()` after loading the package to see your environment.
```

---

# Final Package Structure

```
sePCA/
├── DESCRIPTION
├── NAMESPACE
├── LICENSE.md
├── README.Rmd
├── README.md
├── NEWS.md          # Changelog
├── R/
│   ├── data.R
│   ├── pca.R
│   ├── plotting.R
│   ├── export.R
│   ├── app_ui.R
│   ├── app_server.R
│   ├── run_app.R
│   └── cli_install.R
├── exec/
│   └── sePCA
├── inst/
│   └── app/
│      └── app.R
├── data/
│   └── example_se.rda
├── data-raw/
│   └── example_se.R
├── man/
│   └── *.Rd
├── tests/
│   ├── testthat/
│   │   ├── test-data.R
│   │   ├── test-pca.R
│   │   ├── test-cli_install.R
│   │   └── ...
│   └── testthat.R
├── vignettes/
│   ├── getting-started.Rmd
│   ├── shiny-app.Rmd
│   └── cli.Rmd
├── _pkgdown.yml
└── .github/
    └── workflows/
        ├── R-CMD-check.yaml
        └── pkgdown.yaml
```

---

# Session Info

```{r session-info, eval=TRUE}
sessionInfo()
```
