---
title: "Lecture 5: Lab – R Package Development (devtools)"
author: "Jared Andrews"
date: "April 15, 2026"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  eval = FALSE  # Don't actually run package creation code when knitting
)
```

# Motivation

Many research projects begin as scripts that only make sense to the original author and only run in one working directory. Packaging forces you to turn that private workflow into something installable, documented, and reusable by other people.

This can also personally save you time and sanity in a variety of ways - generic reuseability, minimal editing to apply to new datasets/experiments, preparedness for publication, and not having to remember exactly which script had which bit of code you used in that one analysis 6 months ago.

## Learning Objectives

By the end of this lab, you will be able to:

1. Use `usethis` and `devtools` to scaffold package infrastructure instead of hand-building it
2. Define a small public API and distinguish exported functions from internal helpers
3. Define and validate a correct DESCRIPTION file, including dependencies and metadata
4. Add roxygen2 documentation and generate help files/NAMESPACE
5. Publish the package to a Git repository and install it from GitHub
6. Diagnose and resolve common development issues (namespacing, missing Imports, R CMD check warnings)

---

# Taking the Next Step

In Lecture 4, we created an "analysis core" — a set of functions in a script. But scripts have limitations:

| Script | Package |
|--------|---------|
| `source("path/to/file.R")` | `library(ADS8192)` |
| Paths break when you move files | Installed; works anywhere |
| No formal dependency management | Explicit dependency management |
| No formal help documentation | `?run_pca` works |
| Hard to share | `install_github("you/ADS8192")` |
| No tests to ensure correctness | Automated testing simple to implement |

An R package is barely more complicated than a script, but it provides many benefits even if you're the only one who will ever use it. It can be a little intimating at first, but once you build a package once, it becomes a natural way to organize code.

## Package Development Tools

Fortunately, there are a number of R packages that provide scaffolding and convenience functions to make package development easier:

- `usethis` quickly generates package infrastructure (DESCRIPTION, NAMESPACE, R/ directory, README, data-raw/, etc) and automates common tasks like adding dependencies, creating documentation stubs, and setting up Github
- `devtools` standardizes the edit-document-test-check loop
- `roxygen2` parses specially formatted comments to generate documentation automatically

We'll use all of these tools to help build our package.

# Pre-Lab Checklist

Before we begin, ensure you have:

- [ ] Your `analysis_core.R` script from Lecture 4 with analysis functions
- [ ] Git installed and configured (`git config --global user.name "Your Name"`)
- [ ] A GitHub account

```{r install-dev-tools, eval=FALSE}
# Install development tools if needed
install.packages(c("devtools", "usethis", "roxygen2", "testthat"))
```


---

# Part 1: Create the Package Skeleton

## Step 1: Create the Package

Open RStudio and create a new package:

```{r create-package}
library(usethis)

# Create the package (choose your own name!)
# This creates a new directory with the package structure
create_package("~/ADS8192")  # Or wherever you want it

# This will open a new RStudio session in the package directory
```

After running this, you'll see:

```
ADS8192/
├── .Rbuildignore
├── .gitignore
├── DESCRIPTION
├── NAMESPACE
├── R/
└── ADS8192.Rproj
```

> **Note:** Two important files are created automatically by `usethis`:
>
> - **`.Rbuildignore`** — tells `R CMD build` which files to exclude from the package tarball. For example, the `.Rproj` file, `data-raw/` scripts, and development notes should be listed here. They don't belong in the installed package.
> - **`.gitignore`** — tells Git which files not to track. This typically includes compiled artifacts (`.o`, `.so`), `.Rproj.user/`, and temporary files. You rarely need to edit these manually — `usethis` functions like `use_data_raw()` automatically add the right entries.

## Step 2: Initialize Git

```{r init-git}
# Initialize git repository
use_git()

# This will:
# 1. Create a .git directory
# 2. Make an initial commit
# 3. Potentially restart RStudio
```

## Step 3: Connect to GitHub

```{r use-github}
# Create a GitHub repository and push
use_github()

# This will:
# 1. Create a new repo on GitHub
# 2. Add the remote
# 3. Push your initial commit

# If this doesn't work, you can create the repo manually on GitHub
# and then add the remote via the terminal:
# git remote add origin https://github.com/your-username/ADS8192.git
# git push -u origin main
```

## Step 4: Add a License

```{r add-license}
# MIT License is a good default for open source
use_mit_license()

# This adds:
# - LICENSE.md file
# - License field in DESCRIPTION
```

---

# Part 2: The DESCRIPTION File

The `DESCRIPTION` file is the heart of your package metadata. Open it and edit:

```yaml
Package: ADS8192
Title: ADS 8192 PCA Analysis Tools
Version: 0.0.0.9000
Authors@R:
    person("Your", "Name", email = "you@example.com", role = c("aut", "cre"),
           comment = c(ORCID = "YOUR-ORCID-ID"))
Description: Provides functions to perform principal component analysis (PCA)
    on SummarizedExperiment objects. Includes tools for selecting top variable
    features, running PCA, and creating publication-ready visualizations.
    Designed as a teaching example for scientific application development.
License: MIT + file LICENSE
Encoding: UTF-8
Roxygen: list(markdown = TRUE)
RoxygenNote: 7.2.3
```

## Adding Dependencies

**Critical:** In packages, we don't use `library()`. Instead, we declare dependencies in DESCRIPTION.

```{r add-dependencies}
# Add packages to Imports (required to run)
use_package("SummarizedExperiment")
use_package("ggplot2")
use_package("rlang")  # For .data pronoun in ggplot aes()

# Add packages to Suggests (optional, for examples/tests)
use_package("testthat", type = "Suggests")
use_package("knitr", type = "Suggests")

# NOTE: For Bioconductor packages, you may also need:
# BiocManager::install("SummarizedExperiment")
# and add 'biocViews:' to your DESCRIPTION (can be left empty)
```

After running these, your DESCRIPTION will include:

```yaml
Imports:
    ggplot2,
    SummarizedExperiment
Suggests:
    knitr,
    testthat
```

> **Exercise B:** Why do we put `testthat` in Suggests rather than Imports? What's the difference?

---

# Part 3: Move Functions to R/

## Package Structure for Code

All R code goes in the `R/` directory. You can organize it however you like, but a common pattern:

```
R/
├── data.R         # Functions for data handling (top_variable_features)
├── pca.R          # PCA-related functions
├── plotting.R     # Visualization functions
└── ADS8192-package.R  # Package-level documentation
```

## Create the Files

```{r create-r-files}
# Create R files (this just opens them in RStudio)
use_r("data")      # For top_variable_features()
use_r("pca")       # For run_pca(), pca_variance_explained()
use_r("plotting")  # For plot_pca()
```

## Move and Adapt Your Functions

Copy your functions from `analysis_core.R`, but make these changes:

### Before (Script Style)

```{r script-style, eval=FALSE}
library(SummarizedExperiment)  # DON'T do this in packages!
library(ggplot2)

top_variable_features <- function(se, n = 500, assay_name = "counts") {
  mat <- assay(se, assay_name)  # Uses assay() directly
  # ...
}
```

### After (Package Style)

Two options for using functions from other packages:

**Option 1: Full namespace (recommended for clarity)**

```{r namespace-style, eval=FALSE}
# In R/data.R
top_variable_features <- function(se, n = 500, assay_name = "counts") {
  # Use pkg::fun() syntax
  mat <- SummarizedExperiment::assay(se, assay_name)
  # ...
}
```

**Option 2: Import via roxygen2**

```{r import-style, eval=FALSE}
#' @importFrom SummarizedExperiment assay colData rowData
top_variable_features <- function(se, n = 500, assay_name = "counts") {
  # Can use function directly because it's imported
  mat <- assay(se, assay_name)
  # ...
}
```

### Complete Example: R/data.R

```{r data-r}
# R/data.R

#' Select top variable features
#'
#' @param se A SummarizedExperiment object
#' @param n Number of top variable features to select (default: 500)
#' @param assay_name Name of assay to use (default: "counts")
#'
#' @return A SummarizedExperiment subset to the top n variable features
#' @export
#'
#' @examples
#' # Assuming 'se' is a SummarizedExperiment
#' # se_top <- top_variable_features(se, n = 500)
top_variable_features <- function(se, n = 500, assay_name = "counts") {
    mat <- SummarizedExperiment::assay(se, assay_name)
    vars <- apply(mat, 1, var)
    top_idx <- order(vars, decreasing = TRUE)[seq_len(min(n, length(vars)))]
    se[top_idx, ]
}
```

Note the key additions:

- `#' @export` — Makes the function available to users
- `#' @param` — Documents each parameter
- `#' @return` — Documents the return value
- `#' @examples` — Provides runnable examples
- `pkg::fun()` syntax — Explicit namespacing

---

# Part 4: Generate Documentation

## Run roxygen2

```{r document}
library(devtools)

# Generate documentation and NAMESPACE
document()
```

This creates:

- `man/top_variable_features.Rd` — Help file for each documented function
- `NAMESPACE` — Updated with exports and imports

Check your `NAMESPACE` file:

```
# Generated by roxygen2: do not edit by hand

export(pca_variance_explained)
export(plot_pca)
export(run_pca)
export(top_variable_features)
```

## Test the Help

```{r test-help}
# Load the package in development mode
load_all()

# Test help pages
?top_variable_features
?run_pca
```

---

# Part 5: Add Example Data

Let's include a small example dataset so users (and tests) can try the functions without loading external data.

```{r data-raw}
# Create data-raw directory for data preparation scripts
use_data_raw("example_se")
```

This creates `data-raw/example_se.R`. Edit it:

```{r example-data-script, class.source = "fold-hide"}
# data-raw/example_se.R — creates example_se dataset

# Create a small example SummarizedExperiment
library(SummarizedExperiment)

set.seed(42)

# 100 genes, 8 samples
n_genes <- 100
n_samples <- 8

# Simulate counts (negative binomial-ish)
counts <- matrix(
    rpois(n_genes * n_samples, lambda = 100),
    nrow = n_genes,
    ncol = n_samples
)
rownames(counts) <- paste0("gene", seq_len(n_genes))
colnames(counts) <- paste0("sample", seq_len(n_samples))

# Add some structure: first 20 genes differ by treatment
treatment <- rep(c("control", "treated"), each = 4)
counts[1:20, treatment == "treated"] <- counts[1:20, treatment == "treated"] * 2

# Sample metadata
sample_data <- data.frame(
    sample_id = colnames(counts),
    treatment = treatment,
    batch = rep(c("A", "B"), times = 4),
    row.names = colnames(counts)
)

# Gene metadata
gene_data <- data.frame(
    gene_id = rownames(counts),
    gene_symbol = paste0("SYM", seq_len(n_genes)),
    row.names = rownames(counts)
)

# Create SummarizedExperiment
example_se <- SummarizedExperiment(
    assays = list(counts = counts),
    colData = sample_data,
    rowData = gene_data
)

# Save to data/
usethis::use_data(example_se, overwrite = TRUE)
```

Run the script to create the data:

```{r run-data-script}
source("data-raw/example_se.R")
```

Now document the data:

```{r document-data}
use_r("data-documentation")
```

Add to `R/data-documentation.R`:

```{r data-docs}
#' Example SummarizedExperiment for testing
#'
#' A small SummarizedExperiment with 100 genes and 8 samples.
#' Includes a treatment effect in the first 20 genes.
#'
#' @format A SummarizedExperiment with:
#' \describe{
#'   \item{assays}{counts - raw count matrix}
#'   \item{colData}{sample_id, treatment (control/treated), batch (A/B)}
#'   \item{rowData}{gene_id, gene_symbol}
#' }
#'
#' @source Simulated data for teaching purposes
#'
#' @examples
#' data(example_se)
#' example_se
#' colData(example_se)
"example_se"
```

---

# Part 6: R CMD check

This is the moment of truth! `R CMD check` runs ~50 checks to validate your package.

```{r check}
# Run the full check
check()
```

## Common Issues and Fixes

### 1. Missing Imports

```
Error: could not find function "SummarizedExperiment"
```

**Fix:** Add `#' @importFrom` or use `pkg::fun()` syntax.

### 2. Undocumented Arguments

```
Warning: Undocumented arguments in documentation object 'top_variable_features'
  'row_data'
```

**Fix:** Add `@param row_data` to your roxygen2 block.

### 3. Non-ASCII Characters

```
Warning: found non-ASCII string
```

**Fix:** Replace special characters (curly quotes, em-dashes) with ASCII equivalents.

### 4. Missing Suggests for Examples

```
Package suggested but not available: 'foo'
```

**Fix:** Wrap example code in `\dontrun{}` or add the package to Suggests.

> **Exercise C:** Run `check()` on your package. Categorize each WARNING and NOTE as either "must fix" or "acceptable for now."

---

# Part 7: Push to GitHub and Test Install

## Commit and Push

```{r git-commit}
# Stage all files
# git add -A

# Commit
# git commit -m "Initial package structure with core PCA functions"

# Push
# git push
```

In RStudio, use the Git pane (Ctrl+Alt+M to commit).

## Test Installation

Open a **fresh R session** (important!) and try:

```{r test-install, eval=FALSE}
# Install from GitHub
# remotes::install_github("your-username/ADS8192")

# Load and test
library(ADS8192)
data(example_se)

result <- run_pca(example_se)
plot_pca(result, color_by = "treatment")
```

---

# Part 8: Add a README

```{r readme}
use_readme_rmd()
```

Edit `README.Rmd`:

````markdown
---
output: github_document
---

# ADS8192

<!-- badges: start -->
<!-- badges: end -->

ADS8192 provides tools for performing PCA on SummarizedExperiment objects.

## Installation

```r
# Install from GitHub
remotes::install_github("your-username/ADS8192")
```

## Quick Start

```r
library(ADS8192)
library(airway)

# Load example data
data(airway)

# Run PCA
result <- run_pca(airway, n_top = 500)

# Plot
plot_pca(result, color_by = "dex")
```
````

Then build the README:

```{r build-readme}
build_readme()
```

---

# Summary

Today we:

1. Created a package skeleton with `usethis::create_package()`
2. Set up Git and GitHub for version control
3. Configured DESCRIPTION with proper metadata and dependencies
4. Moved our functions to R/ with roxygen2 documentation
5. Created example data for testing
6. Ran `devtools::check()` and fixed issues
7. Published to GitHub and tested installation

## Package Milestone

✅ A valid GitHub-hosted R package that installs successfully and exposes the analysis core as documented exported functions.

---

## Debrief & Reflection

Before moving on, make sure you can explain:

- Why is `usethis` worth reusing instead of hand-writing package scaffolding?
- Which of your current functions are important for a user to be able to access, and which (if any) could become internal helpers?
- If a collaborator only reads the README and function help, what package design choices will make the code feel coherent instead of script-like?

---

# After-Class Tasks

## Reading

- Skim the Bioconductor Contributions book sections on package structure
- R Packages book: https://r-pkgs.org (chapters 1-7)

## Micro-task 1: Package-Level Documentation

Add a package-level documentation file:

```{r pkg-docs}
use_package_doc()
```

Edit `R/ADS8192-package.R` to describe the package purpose and main functions.

## Micro-task 2: Fresh Install Test

From a fresh R session (restart R first!):

1. Install your package: `remotes::install_github("you/ADS8192")`
2. Run the README example
3. Fix any issues

---

# Quick Reference

```{r quick-ref, eval=FALSE, class.source = "fold-hide"}
# Package development quick reference
library(devtools)
library(usethis)

# Create package
create_package("path/to/pkgname")

# Add infrastructure
use_git()
use_github()
use_mit_license()
use_readme_rmd()

# Add dependencies
use_package("ggplot2")              # Imports
use_package("testthat", "Suggests") # Suggests

# Create files
use_r("filename")                   # R code
use_data_raw("dataname")            # Data prep script

# Development cycle
load_all()      # Load package for testing (Ctrl+Shift+L)
document()      # Generate docs (Ctrl+Shift+D)
check()         # Run R CMD check (Ctrl+Shift+E)
test()          # Run tests (Ctrl+Shift+T)
build_readme()  # Render README.Rmd

# Install
install()       # Install locally
# install_github("user/repo")  # From GitHub
```

---

# Session Info

```{r session-info, eval=TRUE}
sessionInfo()
```
