---
title: "Lecture 9: Lab – CLI Tool Development (R) – CLI Design and Rapp Basics"
author: "Jared Andrews"
date: "April 17, 2026"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  eval = FALSE
)
```

# Motivation

Scientific software often needs to run without a human sitting at the keyboard. Command-line interfaces matter because they make analysis tools usable in pipelines, schedulers, shared compute environments, and repeatable scripts.

This lecture is important for robust software design because a good CLI forces you to make the interface contract explicit: inputs, outputs, defaults, error behavior, and file formats all become visible. That clarity saves time for automation users and reduces drift between the command line and the underlying package logic.

## Learning Objectives

By the end of this session, you will be able to:

1. Implement a simple CLI using Rapp that calls package functions
2. Write usage/help text and define clear inputs/outputs for commands
3. Test CLI functionality from the terminal and interpret exit behavior
4. Decide when a CLI is worth adding and keep it a thin wrapper over package logic
5. Explain why CLIs improve interoperability and pipeline integration

## Scientific Use Case

A bioinformatics core wants to run your analysis across 40 cohorts in a nightly workflow. They do not want a web app and they do not want to source R files by hand. How can this be done in a way that is robust, reproducible, and easy to maintain?

---

# Why Build a CLI?

## The Three Interfaces

We've built:

- **R API**: For users working interactively in R
- **Shiny App**: For point-and-click exploration

Now we add:

- **CLI**: For pipeline integration and non-interactive use

## CLI Use Cases

```bash
# Run analysis in a pipeline
snakemake_rule:
    ADS8192 pca --counts counts.csv --meta samples.csv --output results/

# Process multiple datasets
for dataset in data/*; do
    ADS8192 pca --counts "$dataset/counts.csv" --meta "$dataset/meta.csv" \
        --output "results/$(basename $dataset)/"
done

# Reproducible scripts
#!/bin/bash
ADS8192 pca --counts $1 --meta $2 --n-top 1000 --output pca_results/
```

Benefits:

- **Pipeline integration**: Works with Snakemake, Nextflow, Make
- **Scriptable**: Easy to automate many runs
- **Reproducible**: Command line is self-documenting
- **No IDE required**: Run on servers, HPC clusters

---

## When Should You Add a CLI?

Add a CLI when users need automation, batch execution, scheduler integration, or a stable entry point from another language or workflow manager. Do not add one just because a terminal interface looks advanced. A CLI is worthwhile when it exposes a clear contract around existing package logic.

## Why Use Rapp Instead of Parsing Arguments Yourself?

Argument parsing is infrastructure, not the scientific problem. `Rapp` lets you describe commands, help text, and argument types without hand-writing a parser, which leaves your time for the design work that matters: stable flags, explicit file formats, and DRY delegation to the package core.

# Part 1: CLI Design Principles

## Good CLI Design

### 1. Clear Help Text

```bash
$ ADS8192 pca --help

Usage: ADS8192 pca [OPTIONS]

Run PCA on a SummarizedExperiment and export results.

Options:
  --counts FILE       Path to counts matrix (TSV/CSV) [required]
  --meta FILE         Path to sample metadata (TSV/CSV) [required]
  --output DIR        Output directory [required]
  --n-top INT         Number of top variable genes [default: 500]
  --log-transform     Log-transform counts [default: true]
  --no-log-transform  Don't log-transform counts
  --color-by COL      Metadata column for plot coloring
  -h, --help          Show this message and exit

Examples:
  ADS8192 pca --counts counts.tsv --meta samples.tsv --output results/
  ADS8192 pca --counts counts.tsv --meta samples.tsv --output results/ --n-top 1000
```

### 2. Explicit Inputs and Outputs

Don't read from stdin/write to stdout for data (only for messages):

```bash
# Good: explicit file arguments
ADS8192 pca --counts data.tsv --output results/

# Bad: implicit stdin/stdout
cat data.tsv | ADS8192 pca > results.tsv
```

### 3. Reproducible Defaults

- Document all defaults in help text
- Same inputs → same outputs (deterministic)

### 4. Meaningful Exit Codes

```bash
$ ADS8192 pca --counts missing.tsv --meta meta.tsv --output out/
Error: File not found: missing.tsv

$ echo $?
1  # Non-zero = error
```

| Code | Meaning |
|------|---------|
| 0 | Success |
| 1 | General error |
| 2 | Invalid arguments |

### 5. Machine-Readable Outputs

```bash
# Good: TSV for data
ADS8192 pca ... --output results/
# Creates: results/pca_scores.tsv, results/pca_variance.tsv

# Good: JSON for complex structured data
ADS8192 info --json > info.json
```

---

## Exercise A: Design First

Before implementing, design the help output you want:

```
Usage: ADS8192 pca [OPTIONS]

Run PCA analysis on gene expression data.

Required Arguments:
  --counts FILE      Path to counts matrix (genes × samples, TSV/CSV)
  --meta FILE        Path to sample metadata (TSV/CSV, row names = sample IDs)
  --output DIR       Directory for output files

Optional Arguments:
  --n-top INT        Number of top variable genes [default: 500]
  --log-transform    Log2-transform counts (add pseudocount of 1) [default]
  --no-log-transform Skip log transformation
  --color-by COL     Metadata column for coloring (for plot)
  --assay NAME       Assay name in SE [default: "counts"]
  --help             Show this help message

Outputs:
  {output}/pca_scores.tsv      PCA scores with sample metadata
  {output}/pca_variance.tsv    Variance explained by each PC
  {output}/pca_plot.png        PCA scatter plot (if --color-by specified)

Examples:
  ADS8192 pca --counts counts.tsv --meta samples.tsv --output results/
  ADS8192 pca --counts counts.tsv --meta samples.tsv --output results/ \
      --n-top 1000 --color-by treatment
```

---

# Part 2: Introduction to Rapp

## What is Rapp?

**Rapp** (R application) is a lightweight framework for building command-line tools in R. It provides:

- Argument parsing and type coercion
- Help text generation (including `--help-yaml`)
- Subcommand support
- A clean path from script -> CLI

Alternative options include `optparse`, `argparse`, and `docopt`, but in this course we will use Rapp.

## Installation

```{r install-rapp}
# Install Rapp
install.packages("Rapp")

# Or from GitHub for the latest version
# remotes::install_github("r-lib/Rapp")
```

## How Rapp Declares a CLI

Rapp infers CLI structure from normal R code:

- `n_top <- 500L` becomes `--n-top 500`
- `log_transform <- TRUE` becomes `--log-transform` / `--no-log-transform`
- `path <- NULL` becomes a positional argument
- `switch("", cmd1 = { ... }, cmd2 = { ... })` declares commands

You can add help metadata using `#|` annotations:

```r
#| description: Number of top variable genes
#| short: n
n_top <- 500L
```

**Note:** snake_case names automatically map to kebab-case flags (`n_top` -> `--n-top`).

Rapp provides built-in help:

- `--help` shows usage and options
- `--help-yaml` prints machine-readable metadata

## Rapp Variable → CLI Flag Mapping

Rapp infers CLI flags directly from how you declare variables in R. Here is a complete reference:

| R Declaration | CLI Flag | Type | Notes |
|---------------|----------|------|-------|
| `n_top <- 500L` | `--n-top 500` | Integer | `L` suffix = integer; snake_case → kebab-case |
| `log_transform <- TRUE` | `--log-transform` / `--no-log-transform` | Boolean | Logical creates a flag pair |
| `color_by <- ""` | `--color-by VALUE` | String | Empty string = required or optional string arg |
| `counts <- ""` | `--counts FILE` | String | Provide a description annotation for clarity |
| `scale <- 1.5` | `--scale 1.5` | Double | Plain numeric = double |
| `path <- NULL` | positional arg | Any | `NULL` default = positional (no `--` prefix) |

> **Key rule:** snake_case variable names automatically become kebab-case flags. `n_top` → `--n-top`, `log_transform` → `--log-transform`. Do NOT manually rename variables to match your desired flag name — let Rapp handle the conversion.

## Rapp Script Structure and Workflow

A complete Rapp script has this overall structure:

```{r rapp-structure, eval=FALSE}
#!/usr/bin/env Rapp         # Shebang: marks this as a Rapp executable
#| name: ADS8192            # CLI name (used in --help output)
#| title: ADS8192 PCA Tool  # Short title
#| description: ...         # Longer description shown in --help

# Package loading (runs before any command)
suppressPackageStartupMessages({
    library(ADS8192)
    library(utils)
    library(stats)
})

# Optional helper functions (not exposed as CLI args)
read_data_file <- function(path) { ... }

# Global arguments (available to ALL subcommands)
# (declare them here if needed, otherwise omit)

switch(
    "",           # Rapp replaces this "" with the subcommand name at runtime

    #| title: Subcommand title
    #| description: What this subcommand does
    my_command = {
        # Arguments for this subcommand
        #| description: Input file path
        input <- ""

        #| description: Number of features
        n_top <- 500L

        # Command logic here
        result <- run_pca(...)
    },

    another_command = {
        # Arguments for another subcommand
        ...
    }
)
```

Key structural points:
- **Shebang** (`#!/usr/bin/env Rapp`): makes the file directly executable on Unix/Mac after `chmod +x`
- **`switch("")`**: the Rapp idiom for declaring subcommands — Rapp substitutes the actual subcommand name at runtime
- **Variable declarations inside each `case`**: become the subcommand's CLI arguments
- **`#|` annotations**: parsed as YAML by Rapp; provide help text, short flags, etc.
- **Type coercion**: Rapp reads command-line strings and converts them to the declared R type (`500L` → integer, `TRUE` → logical, `""` → character)

## Rapp vs Other CLI Packages in R

Several R packages support building CLIs. Here is a brief comparison:

| Package | Approach | Pros | Cons |
|---------|----------|------|------|
| **Rapp** | Infers from normal R variable declarations + `#\|` annotations | Minimal boilerplate; integrates naturally with package code | Newer, fewer ecosystem examples |
| **optparse** | Explicit `make_option()` calls in a list | Widely used, well-documented, very explicit | Verbose; must declare everything manually |
| **argparse** | Python-argparse style via reticulate | Powerful, familiar to Python users | Requires Python; heavier dependency |
| **docopt** | Parses a usage string you write by hand | Elegant for simple CLIs; self-documenting | Usage string can be brittle; less flexible for complex CLIs |

In this course we use **Rapp** because it lets you describe a CLI with minimal overhead and integrates naturally with the package architecture we've been building. The core design principle — declare variables, add annotations, let the framework generate the parser — keeps CLI code DRY and close to the R functions it wraps.

---

# Part 3: Implementing the CLI

## Step 1: Create the Rapp App in `exec/`

The full CLI script lives at `exec/ADS8192`. Let's walk through it in three parts to understand its structure.

```{r rapp-header, eval=FALSE}
# exec/ADS8192 — Part 1: Header, package loading, and helper functions
#!/usr/bin/env Rapp
#| name: ADS8192
#| title: ADS8192 PCA Tool
#| description: PCA analysis for SummarizedExperiment data.

suppressPackageStartupMessages({
    library(ADS8192)
    library(utils)
    library(stats)
})

# Helper to read TSV/CSV (not exported; kept in CLI script)
read_data_file <- function(path) {
    ext <- tolower(tools::file_ext(path))
    if (ext == "csv") {
        read.csv(path, row.names = 1, check.names = FALSE)
    } else {
        utils::read.table(path, sep = "\t", header = TRUE, row.names = 1,
                          check.names = FALSE)
    }
}
```

The `pca` subcommand declares its arguments as R variables, validates inputs, reads files, calls the package core functions, and writes output:

```{r rapp-pca-cmd, class.source = "fold-hide", eval=FALSE}
# exec/ADS8192 — Part 2: pca subcommand
switch(
    "",

    #| title: Run PCA analysis
    #| description: Run PCA on a counts matrix and sample metadata, export results.
    pca = {
        #| description: Path to counts matrix (TSV/CSV, genes x samples)
        #| short: c
        counts <- ""

        #| description: Path to sample metadata (TSV/CSV)
        #| short: m
        meta <- ""

        #| description: Output directory
        #| short: o
        output <- ""

        #| description: Number of top variable genes
        #| short: n
        n_top <- 500L

        #| description: Log-transform counts
        log_transform <- TRUE

        #| description: Metadata column for plot coloring (optional)
        color_by <- ""

        # Validation
        if (counts == "" || meta == "" || output == "") {
            stop("--counts, --meta, and --output are required", call. = FALSE)
        }
        if (!file.exists(counts)) stop("File not found: ", counts, call. = FALSE)
        if (!file.exists(meta)) stop("File not found: ", meta, call. = FALSE)

        if (!dir.exists(output)) dir.create(output, recursive = TRUE)

        # Read inputs
        counts_df <- read_data_file(counts)
        meta_df <- read_data_file(meta)

        # Build SE and run analysis using package core functions
        se <- SummarizedExperiment::SummarizedExperiment(
            assays = list(counts = as.matrix(counts_df)),
            colData = meta_df
        )
        result <- run_pca(se, n_top = n_top, log_transform = log_transform)

        save_pca_results(result, output)

        if (color_by != "") {
            plot_file <- file.path(output, "pca_plot.png")
            p <- plot_pca(result, color_by = color_by)
            ggplot2::ggsave(plot_file, p, width = 8, height = 6, dpi = 150)
            message("Saved: ", plot_file)
        }

        message("Done.")
    },
```

The `validate` subcommand checks inputs without running PCA — useful for pipeline pre-flight checks:

```{r rapp-validate-cmd, class.source = "fold-hide", eval=FALSE}
# exec/ADS8192 — Part 3: validate subcommand
    #| title: Validate input files
    #| description: Check that input files exist, parse correctly, and report dimensions.
    validate = {
        #| description: Path to counts matrix (TSV/CSV)
        #| short: c
        counts <- ""

        #| description: Path to sample metadata (TSV/CSV)
        #| short: m
        meta <- ""

        if (counts == "" || meta == "") {
            stop("--counts and --meta are required", call. = FALSE)
        }
        if (!file.exists(counts)) stop("File not found: ", counts, call. = FALSE)
        if (!file.exists(meta)) stop("File not found: ", meta, call. = FALSE)

        counts_df <- read_data_file(counts)
        meta_df <- read_data_file(meta)

        message("Counts dimensions: ", nrow(counts_df), " genes x ",
                ncol(counts_df), " samples")
        message("Metadata rows: ", nrow(meta_df))
        message("Metadata columns: ",
                paste(colnames(meta_df), collapse = ", "))

        if (!all(colnames(counts_df) %in% rownames(meta_df))) {
            stop("Sample IDs in counts do not match metadata row names",
                 call. = FALSE)
        }

        message("All sample IDs match. Inputs look valid.")
    }
)
```

> **Note:** The `switch("")` structure spans blocks 2 and 3 — in the actual `exec/ADS8192` file, the `validate` case is inside the same `switch()` call as `pca`, and the closing `)` is at the very end of the file. The three sections above show the logical structure.

## Step 2: Run During Development

```bash
# Show top-level help
Rapp exec/ADS8192 --help

# Command-specific help
Rapp exec/ADS8192 pca --help

# Run PCA
Rapp exec/ADS8192 pca     --counts tests/testdata/counts.tsv     --meta tests/testdata/meta.tsv     --output tests/testdata/output/
```

## Step 3: Interactive Testing from R

```r
# Run from an R session (useful for debugging)
Rapp::run("exec/ADS8192", c(
    "pca",
    "--counts", "tests/testdata/counts.tsv",
    "--meta", "tests/testdata/meta.tsv",
    "--output", "tests/testdata/output/",
    "--n-top", "500"
))
```

---

## Step 4: Export a Launcher Installer

Running `Rapp exec/ADS8192 pca --help` works during development, but after a
user installs your package from GitHub, the `exec/` directory is buried inside
the R library tree. Rapp solves this with **launchers** — lightweight
shell scripts (`.bat` on Windows) that live on `PATH` and forward to the
installed Rapp app.

### How It Works

```r
# One-liner: install launchers for every Rapp in a package's exec/
Rapp::install_pkg_cli_apps("ADS8192")
```

After this, users can invoke the CLI directly:

```bash
ADS8192 pca --help
ADS8192 pca --counts counts.tsv --meta meta.tsv --output results/
```

On Windows, `install_pkg_cli_apps()` creates `.bat` wrappers in
`%LOCALAPPDATA%\Programs\R\Rapp\bin` and adds that directory to `PATH`.
On macOS / Linux, launchers land in `~/.local/bin` (usually already on `PATH`).

### Export a Thin Wrapper

Rather than asking users to remember the `Rapp::install_pkg_cli_apps()` call,
export your own convenience function. Create `R/install_cli.R`:

```r
#' Install ADS8192 CLI launchers
#'
#' Places lightweight launcher scripts on the user's `PATH` so the
#' ADS8192 CLI can be invoked directly from a terminal (e.g. `ADS8192 pca --help`).
#'
#' @inheritDotParams Rapp::install_pkg_cli_apps -package -lib.loc
#' @export
install_ADS8192_cli <- function(...) {
    Rapp::install_pkg_cli_apps(package = "ADS8192", lib.loc = NULL, ...)
}
```

Then document it, `devtools::document()`, and users get:

```r
remotes::install_github("you/ADS8192")
ADS8192::install_ADS8192_cli()
```

…followed by seamless terminal usage.

### Try It Now

```bash
# Install launchers for ADS8192 (run once after installation)
Rscript -e "Rapp::install_pkg_cli_apps('ADS8192')"

# Now use the CLI directly
ADS8192 --help
ADS8192 pca --help
```

Add the install instructions to your README so users know about it.

---

# Part 4: Testing the CLI

## Manual Testing

### Create Test Data

```{r create-test-data}
# Create small test files
library(SummarizedExperiment)
library(airway)
data(airway)

# Extract and save counts
counts <- assay(airway, "counts")[1:1000, ]  # First 1000 genes
write.table(
    counts,
    "tests/testdata/counts.tsv",
    sep = "\t",
    quote = FALSE
)

# Save metadata
meta <- as.data.frame(colData(airway))
write.table(
    meta,
    "tests/testdata/meta.tsv",
    sep = "\t",
    quote = FALSE
)
```

### Run from Terminal

```bash
# Navigate to package directory
cd ADS8192

# Test help
Rapp exec/ADS8192 --help

# Test PCA command help
Rapp exec/ADS8192 pca --help

# Run PCA
Rapp exec/ADS8192 pca \
    --counts tests/testdata/counts.tsv \
    --meta tests/testdata/meta.tsv \
    --output tests/testdata/output/

# Check outputs
ls tests/testdata/output/
cat tests/testdata/output/pca_variance.tsv

# Test with plot
Rapp exec/ADS8192 pca \
    --counts tests/testdata/counts.tsv \
    --meta tests/testdata/meta.tsv \
    --output tests/testdata/output2/ \
    --color-by dex
```

## Testing Error Handling

```bash
# Missing required argument
Rapp exec/ADS8192 pca --counts counts.tsv
# Should error: --counts, --meta, and --output are required

# Missing file
Rapp exec/ADS8192 pca \
    --counts nonexistent.tsv \
    --meta meta.tsv \
    --output out/
# Should error: File not found: nonexistent.tsv

# Check exit code
Rapp exec/ADS8192 pca --counts nonexistent.tsv --meta meta.tsv --output out/
echo $?  # Should be 1
```

---

## Exercise B: Enhance Validate Command

Improve the `validate` subcommand so it:

1. Checks if input files exist
2. Checks if they can be parsed
3. Reports dimensions
4. Checks for sample ID matches

Example output:

```bash
$ Rapp exec/ADS8192 validate --counts counts.tsv --meta meta.tsv

Validating counts file: counts.tsv
  ✓ File exists
  ✓ Can be parsed
  ✓ Dimensions: 10000 genes × 8 samples

Validating metadata file: meta.tsv
  ✓ File exists
  ✓ Can be parsed
  ✓ Rows: 8

Checking consistency:
  ✓ Sample IDs match
  ✓ Metadata columns: cell, dex, albut, Run, avgLength, Experiment, Sample, BioSample

Ready for analysis!
```

---

## Exercise C: DRY Check

Verify that your CLI calls `run_pca()` from your package rather than re-implementing the PCA logic.

Look for:

- `run_pca(...)` call inside the `pca` command in `exec/ADS8192`
- `plot_pca(...)` call for plotting when `--color-by` is set
- No direct calls to `prcomp()` in the CLI script

---

# Part 5: Versioning and Stability

## The Contract

CLI users depend on:

- **Argument names**: `--n-top` shouldn't become `--ntop`
- **Output file names**: `pca_scores.tsv` shouldn't become `scores.tsv`
- **Output format**: TSV columns shouldn't change unexpectedly

## Preventing Common Formatting Errors

### Argument Names

Rapp converts snake_case to kebab-case automatically. Never manually rename a variable to match the flag name you want:

```{r arg-names, eval=FALSE}
# CORRECT: let Rapp handle the conversion
n_top <- 500L       # → --n-top
log_transform <- TRUE  # → --log-transform / --no-log-transform
color_by <- ""      # → --color-by

# WRONG: manually using a different variable name to get a different flag
nTop <- 500L        # → --nTop (camelCase not converted to kebab-case by Rapp)
```

If you need a specific flag name, use the snake_case form that maps to it. `--n-top` comes from `n_top`; `--ntop` would come from `ntop`. Pick the variable name, and the flag name follows.

### Output File Names

Always use literal string names for output files — never derive them from user input or parameter values:

```{r output-names, eval=FALSE}
# CORRECT: literal, stable names
scores_file <- file.path(output, "pca_scores.tsv")
var_file    <- file.path(output, "pca_variance.tsv")

# WRONG: names derived from inputs
scores_file <- file.path(output, paste0(prefix, "_scores.tsv"))  # prefix could be anything
```

Stable output names let pipeline users hard-code the expected files. Changing `pca_scores.tsv` to `scores.tsv` in a future version is a breaking change.

### Output Format and Column Names

Column names in output files are part of the user contract. If `result$scores` has columns `PC1`, `PC2`, ..., `sample_id`, `treatment`, those column names are what pipeline users depend on.

- **Adding new columns**: generally backward-compatible (minor version bump)
- **Renaming existing columns**: breaking change (major version bump)
- **Changing column order**: avoid it; use column names in downstream code, not positional indexing

### Testing for Stability

Use snapshot tests to protect output structure:

```{r snapshot-test, eval=FALSE}
# tests/testthat/test-cli-stability.R
test_that("pca output columns are stable", {
    data(example_se, package = "ADS8192")
    result <- run_pca(example_se, n_top = 50)

    # Snapshot the column names — this will fail if columns change
    expect_snapshot(colnames(result$scores))
    expect_snapshot(colnames(pca_variance_explained(result)))
})
```

When you intentionally change output structure, update the snapshot with `testthat::snapshot_review()`.

## Semantic Versioning

- **MAJOR** (1.0.0 → 2.0.0): Breaking changes to CLI interface
- **MINOR** (1.0.0 → 1.1.0): New features, backward compatible
- **PATCH** (1.0.0 → 1.0.1): Bug fixes

## What's Breaking?

| Change | Breaking? |
|--------|-----------|
| Rename `--n-top` to `--n_top` | Yes |
| Add new optional flag | No |
| Remove a flag | Yes |
| Change output column names | Yes |
| Add new output columns | Usually no |
| Change default value | Maybe |

Document changes in a CHANGELOG!

---

# Summary

Today we:

1. Learned CLI design principles (help text, explicit I/O, exit codes)
2. Built a CLI using Rapp with commands and options
3. Implemented the `pca` command that calls our package functions
4. Exported a launcher installer (`install_ADS8192_cli()`) for seamless terminal use
5. Tested error handling and exit codes
6. Discussed versioning and stability

## Package Milestone

✅ A working CLI prototype with an exported launcher installer — users can run
`ADS8192 pca --help` directly from the terminal after installing your package.

---

## Debrief & Reflection

Before moving on, make sure you can answer:

- When is a CLI genuinely useful, and when is it just an unnecessary extra interface?
- Which parts of the CLI are stable user contract, and which parts are replaceable implementation details?
- How does `Rapp` let you avoid reinventing plumbing so you can focus on I/O contracts and DRY reuse of the package core?

---

# After-Class Tasks

## Micro-task 1: Add a Flag

Add one additional CLI flag (e.g., `--pcs 1,2` or `--n-top 500`) and ensure it changes outputs appropriately.

## Micro-task 2: Document CLI + Launcher in README

Add a "Command Line Interface" section to README with:

- Launcher installation command (`yourpkg::install_yourpkg_cli()`)
- Basic usage example using the launcher (e.g., `ADS8192 pca --help`)
- List of available commands
- Fallback instructions (`Rapp exec/ADS8192 pca --help`) for users who skip the launcher

## Micro-task 3: pkgdown Article

Add a vignette/article for CLI usage:

```{r cli-vignette}
usethis::use_vignette("cli", title = "Command Line Interface")
```

---

# CLI Quick Reference

```bash
# Install launchers (one-time, after package installation)
Rscript -e "ADS8192::install_ADS8192_cli()"
# Or equivalently:
Rscript -e "Rapp::install_pkg_cli_apps('ADS8192')"

# After installing launchers — the primary way to use the CLI
ADS8192 --help
ADS8192 pca --help
ADS8192 pca \
    --counts counts.tsv \
    --meta samples.tsv \
    --output results/ \
    --n-top 500 \
    --color-by treatment

# During development (before launchers are installed)
Rapp exec/ADS8192 --help
Rapp exec/ADS8192 pca --help

# Check exit code
echo $?

# Redirect stderr (messages) vs stdout (none in our design)
ADS8192 pca ... 2> log.txt
```

---

# Session Info

```{r session-info, eval=TRUE}
sessionInfo()
```
