---
title: "Lecture 10: GitHub as a Collaboration Platform"
author: "Jared Andrews"
date: "April 17, 2026"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  eval = FALSE
)
```

# Motivation

Git tracks history. GitHub adds tooling around that history: a place to discuss what should change, propose a change, review it, and remember why it happened after the fact.

Earlier lectures already had you push code to GitHub, wire up CI, and publish a pkgdown site. Those are the mechanical parts. This session is about the collaboration features built on top: **issues**, **pull requests**, and **code review**. Even if you never work on a team, these features are some of the most useful things GitHub offers a solo developer. Future-you is a collaborator too.

## Learning Objectives

By the end of this session, you will be able to:

1. Explain what issues, pull requests, and code review are for, and when each is worth the overhead
2. Use these features on your own package to keep notes, propose changes, and self-review before merging
3. Participate in a code review as either author or reviewer with reasonable norms
4. Decide when a lightweight process (direct commit) is appropriate versus when a heavier process (branch + PR + review) is worth it

## Scientific Use Case

A collaborator tries your package and their CLI run crashes on a TSV with an empty column. They file an issue with the exact command and the error. You reproduce it on a branch, fix the input validation, and open a pull request. A labmate reviews the diff and notices that your fix *also* silently changed the default behavior when the column is present but all-zero. You split that second change into its own issue rather than sneaking it into the bugfix. The bug is fixed, the behavior change is discussed openly, and the history records why both happened.

Every piece of that workflow - the filed issue, the branch, the PR, the review, the split - exists because of GitHub's collaboration features, not Git itself.

---

# Where We Are

By this point in the course you have:

- A valid R package on GitHub with documentation, examples, and tests
- CI (GitHub Actions) running `R CMD check` and deploying a pkgdown site on pushes
- A CLI included in the package that is documented and easily installed

What you **haven't** done yet is use the GitHub features that exist around the code: issues, branches as proposals, pull requests, and review. That's the gap this lecture fills.

---

# Part 1: Issues — The Project's Shared Memory

An **issue** is generally a single-purpose thread: one bug, one feature request, one question, one design note. It has a title, a description, optional labels, optional assignees, and a conversation.

The useful mental model is that issues are the project's shared memory. Anything worth remembering that isn't code belongs in one.

## What Belongs in an Issue

- **Bugs** — what you did, what happened, what you expected, environment.
- **Feature requests** — what you want and (ideally) why.
- **Questions** — "how should this behave when the input has no samples?"
- **Design notes** — "we should think about whether `run_pca()` should center by default."

A good issue title reads like a sentence you could put on a TODO list. "Bug" is a bad title; "`run_pca()` errors on all-zero rows" is a good one.

## Benefit for Solo Work

Using issues can also be useful for solo work.On your own package, issues replace three worse things:

- TODO comments in code (invisible once the file is closed)
- Sticky notes and text files (unsearchable, ungrouped)
- Your memory (unreliable at 2am)

Every time you notice "huh, I should fix that someday," file an issue in 30 seconds. A month later you have a searchable, dated list of everything-you-noticed, potentially grouped, linked to the specific code in question if you so desire.

## Benefit for Teams

The benefits for teams are many. Two people won't unknowingly work on the same bug if it is assigned. A new collaborator can scan open issues to get a sense of what the project is wrestling with. A maintainer triaging a burst of bug reports can close duplicates and merge threads.

## Cross-Linking

Mentioning `#123` anywhere on GitHub — a commit message, a PR description, another issue — creates a link. Writing `Fixes #123` in a pull request description automatically closes the issue when the PR merges. Use this liberally; it is how you turn issues from a pile of notes into a connected history.

An example PR description that links an issue (and automatically closes it when merged):

```
Fix input validation for all-zero rows

Filters zero-variance features before PCA and adds a test.

Fixes #47.
```

This would let you track both the issue and the resolution from either side as they'll be linked.

---

# Part 2: Branches and Pull Requests — Proposing Change, Not Just Pushing It

A **pull request** is a proposal to merge one branch into another. Mechanically it's a diff, but collaboratively, it's a place to discuss that diff before it lands.

## Why Branch + PR Is Better Than Committing to `main`

Even alone:

- **Reversible.** A branch can be abandoned or force-pushed freely. `main` can't.
- **Reviewable.** The PR page is a diff viewer that scales better than `git diff` on 40 files.
- **CI-gated.** Your GitHub Actions workflow (from Lecture 6) runs on the PR. You find out about a broken test *before* `main` is broken, not after.
- **Linked.** The PR ties the change to the issue it resolves and to any follow-up discussion.

## Anatomy of a Good PR

- **Scoped.** One logical change per PR. A bugfix and a refactor should be two PRs.
- **Titled like a commit.** Short, imperative, specific. "Fix crash on all-zero rows" beats "bugfix".
- **Description says *why*.** The diff shows *what*. The description is where you explain the motivation, the alternatives you considered, and anything a reviewer should watch for.
- **Links the issue.** `Fixes #47` at the bottom.

## Draft PRs

Open a **draft PR** when you want CI to run and the conversation to start but you aren't asking for a merge yet. It's the right place for "does this approach look reasonable?" questions — earlier is cheaper than later.

## CI as the First Reviewer

Your `R-CMD-check.yaml` workflow runs automatically on every PR. Treat it as an automated first reviewer: if CI is red, the human reviewers shouldn't have to look yet. Fix it, push again, let CI settle, then ask for review.

---

# Part 3: Code Review — The Highest-Leverage Habit

A **code review** is one person reading another person's proposed diff and leaving comments. On GitHub this happens on the PR's "Files changed" tab.

Review is the single highest-leverage habit in collaborative software. It isn't just about catching bugs.

## What Reviewers Actually Look For

Roughly in order:

1. **Correctness.** Does it do what the description claims? Are there obvious edge cases it misses?
2. **Scope.** Does the diff match the stated goal, or is there extra stuff hiding in it?
3. **Clarity.** Will someone reading this in six months understand it without the author present?
4. **Tests.** Is there enough coverage that a future change would notice if it broke this?
5. **Surprises.** Anything a user of this function would not expect — silent behavior changes, new dependencies, changed defaults.

Style and formatting are the *least* important category. A good reviewer flags blockers, not nits.

## How to Give a Good Review

- **Ask before demanding.** "Is there a reason we aren't using `X` here?" beats "use `X`."
- **Separate blockers from suggestions.** Prefix non-blocking comments with "nit:" or "optional:" so the author knows what must change before merge.
- **Be specific.** "This is confusing" is not useful; "I expected `n_top` to default to all features, not 500" is.
- **Approve when the remaining items are nits.** Don't hold up a merge over taste.

## How to Receive a Good Review

- **Assume good faith.** The reviewer is trying to help, not attack your code.
- **Push follow-up commits.** Don't squash away the review conversation mid-flight.
- **Resolve threads you address.** Leave threads you disagree with open and reply with why.
- **Separate follow-ups into new issues.** "Good point, that's a bigger change — filed as #58" is a legitimate response.

## Benefits Beyond Bug Catching

Bugs caught is the easiest benefit to see, but not the biggest. The deeper wins:

- **Knowledge diffusion.** After a review, at least two people understand that code.
- **Onboarding.** Reviewing is how new contributors learn the codebase without being handed an artificial starter task.
- **Shared ownership.** Code reviewed by the team belongs to the team.
- **Durable rationale.** The review conversation is attached to the diff forever — future archaeologists reading `git blame` can click through and see *why*.

## Self-Review: The Solo Dev's Version

On your own repo, open the PR, then read the diff yourself before merging. Pretend you didn't write it. You'll find a surprising number of leftover `print()` calls, half-renamed variables, and changes you didn't mean to commit. Self-review catches a meaningful fraction of what a second reviewer would, at zero coordination cost.

---

# Part 4: Supporting Features, Briefly

These aren't the main event, but they're worth knowing exist.

**Protected `main` branch.** A repo setting that requires PRs (not direct pushes) and optionally requires passing CI and approving reviews before merge. Turn this on as soon as you have even one collaborator.

**CODEOWNERS.** A file that auto-requests reviews from specific people when paths they own are touched. Useful once the project is big enough that "who should review this?" is a non-obvious question.

**Releases and tags.** A GitHub release is a human-readable wrapper around a Git tag. Tagged releases are the stable handle collaborators install — `remotes::install_github("you/ADS8192@v0.1.0")` pins to a tag, not a moving branch. Write release notes that list what changed; pipeline users scan them before upgrading.

**Discussions.** A forum-style space for open-ended questions that aren't bugs or feature requests. Keeps the issue tracker focused on actionable work.

**Projects.** Kanban-style boards over issues and PRs. Useful when you have enough in flight that a list view stops helping.

---

# Part 5: When to Use What

Process should match stakes. Use the lightest workflow that gives you the benefits you need.

| Situation | Workflow |
|---|---|
| Typo or one-line doc fix on your own repo | Direct commit to `main` is fine |
| Real change on your own repo | Branch + PR + self-review + merge |
| Collaborator contribution | Issue first, then branch + PR + review + merge |
| Anything touching a released interface | PR + review, always — even solo |
| Exploratory work nobody should depend on | Branch, no PR needed until you want feedback |

The mistake in both directions is worth avoiding. Over-processing a one-line typo is theater. Under-processing a change that ships to users is how silent behavior changes leak into releases.

---

# Summary

Today we covered, at a high level:

1. **Issues** as the project's shared memory — useful solo, indispensable on a team
2. **Branches and pull requests** as proposals, with CI as an automated first reviewer
3. **Code review** as the highest-leverage collaboration habit, including self-review for solo work
4. **Supporting features** — protected branches, CODEOWNERS, releases, discussions, projects
5. **Matching process to stakes** — neither over- nor under-engineering the workflow

None of this is technically hard. It's a set of norms, and the payoff is that your code stays understandable by more than one person and your project's history stays legible after you stop paying attention to it.

---

## Debrief & Reflection

Before moving on, make sure you can answer:

- Why do issues matter on a project where you are the only contributor?
- What does a pull request give you that a direct commit to `main` does not?
- What is the biggest benefit of code review, and why isn't it "catching bugs"?

---

# Session Info

```{r session-info, eval=TRUE}
sessionInfo()
```
