---
title: "Lecture 8: Lab – Shiny (R) – Packaging, Documentation, and Deployment"
author: "Jared Andrews"
date: "April 16, 2026"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  eval = FALSE
)
```

# Learning Objectives

By the end of this session, you will be able to:

1. Add a Shiny application to the package as a function and document it
2. Deploy the application to a hosting platform (shinyapps.io or Posit Cloud)
3. Add inputs for flexibility and interactive elements in outputs
4. Explain why packaging, optional dependencies, and deployment are reproducibility decisions as much as UI decisions
5. Discuss and (optionally) implement a basic testing strategy for Shiny apps

**Course Learning Outcomes (CLOs):** CLO 1, 4, 5, 6

## Motivation

An app is only useful scientific software if other people can install it, launch it, and trust that it behaves the same way outside your own machine. Packaging and deployment turn an interesting interface into a reproducible tool.

This lecture matters because deployment decisions shape maintainability. Stable entry points, optional dependencies, and clean file lookup rules save time for collaborators, reduce environment-specific failures, and keep the app aligned with the same package core used everywhere else.

## Evaluation Checklist

Before packaging or deploying an app, ask:

- Is the app truly a thin layer over tested package functions?
- Which dependencies are essential, and which should stay optional?
- Can a user install and launch the app from a clean session?
- What interface failures need friendly validation or graceful fallback?
- Does the deployment choice fit the intended audience and maintenance model?
- Would packaging this app reduce manual setup more than another custom distribution method?

## Scientific Use Case

Your lab wants to share a small QC dashboard with collaborators at another institution. Some users will explore the app interactively, while others only want the package functions in scripts. How do you package the app so both audiences are supported without forcing every user into the same workflow?

---

# Why Package Your Shiny App?

## Current State: Standalone App

Right now, your Shiny app is a standalone `app.R` file. To run it, users need to:

1. Clone your repo (or copy the file)
2. Install all dependencies manually
3. Run `shiny::runApp("path/to/app.R")`

## Goal: Packaged App

After packaging, users can:

```{r packaged-usage, eval=FALSE}
# Install once
remotes::install_github("you/sePCA")

# Run anywhere
sePCA::run_app()
```

Benefits:

- **Single install command** handles all dependencies
- **Works anywhere** — no file paths to manage  
- **Version-controlled** — users get consistent behavior
- **Documented** — `?run_app` shows how to use it

---

## Why Reuse Package Conventions and Deployment Platforms?

Package conventions such as `inst/`, `system.file()`, and `run_app()` exist so you do not have to invent a custom file layout or startup story for every app. Hosting platforms then give you a managed way to share that app instead of building deployment plumbing yourself.

Reusing those conventions keeps the design problem focused where it belongs: what should the app expose, how should dependencies be declared, and how should failures be handled for real users.

# Part 1: Package Structure for Shiny Apps

## Where Does the App Go?

The conventional location is `inst/app/`:

```
sePCA/
├── DESCRIPTION
├── NAMESPACE
├── R/
│   ├── run_app.R        # Exported function to launch app
│   ├── app_ui.R         # UI definition
│   ├── app_server.R     # Server logic
│   └── mod_*.R          # Module files
├── inst/
│   └── app/
│       └── app.R        # App entry point (calls run_app)
└── ...
```

Why `inst/`?

- Files in `inst/` are copied as-is when the package is installed
- They're accessible via `system.file("app", package = "sePCA")`
- The actual app logic lives in `R/` for documentation and testing

## Step 1: Create the Directory Structure

```{r create-dirs}
# In your package directory
dir.create("inst/app", recursive = TRUE)
```

## Step 2: Create App Files in R/

### R/app_ui.R

```{r app-ui}
# R/app_ui.R

#' Shiny App UI
#'
#' @return A Shiny UI definition
#' @noRd
app_ui <- function() {
    bslib::page_sidebar(
        title = "sePCA Explorer",
        theme = bslib::bs_theme(bootswatch = "flatly"),
        
        sidebar = bslib::sidebar(
            shiny::h4(shiny::icon("cogs"), "Analysis Settings"),
            
            shiny::numericInput(
                "n_top",
                "Top variable genes:",
                value = 500, min = 50, max = 5000, step = 50
            ),
            shiny::checkboxInput("log_transform", "Log-transform counts", TRUE),
            shiny::checkboxInput("scale", "Scale features", TRUE),
            
            shiny::hr(),
            shiny::h4(shiny::icon("palette"), "Visualization"),
            
            shiny::selectInput("color_by", "Color by:", choices = NULL),
            shiny::selectInput("shape_by", "Shape by:", choices = NULL),
            shiny::sliderInput("point_size", "Point size:", 4, 1, 10, 1),
            
            shiny::hr(),
            shiny::downloadButton("download_plot", "Download Plot")
        ),
        
        bslib::navset_card_tab(
            bslib::nav_panel(
                "PCA Plot",
                shiny::plotOutput("pca_plot", height = "500px")
            ),
            bslib::nav_panel(
                "Variance",
                shiny::plotOutput("variance_plot", height = "400px")
            ),
            bslib::nav_panel(
                "Sample Data",
                DT::dataTableOutput("scores_table")
            )
        )
    )
}
```

### R/app_server.R

```{r app-server}
# R/app_server.R

#' Shiny App Server
#'
#' @param input Shiny input
#' @param output Shiny output
#' @param session Shiny session
#'
#' @return NULL (side effects only)
#' @noRd
#'
#' @importFrom SummarizedExperiment colData
#' @importFrom ggplot2 ggplot aes geom_col geom_text theme_minimal labs ylim
#' @importFrom rlang .data
app_server <- function(input, output, session) {
    # Load example data
    # In a real app, you might want to let users upload their own data
    se_data <- shiny::reactive({
        data("airway", package = "airway", envir = environment())
        get("airway", envir = environment())
    })
    
    # Update select inputs based on available metadata
    shiny::observe({
        se <- se_data()
        cols <- colnames(SummarizedExperiment::colData(se))
        shiny::updateSelectInput(session, "color_by", choices = cols)
        shiny::updateSelectInput(session, "shape_by", choices = c("None", cols))
    })
    
    # Compute PCA
    pca_result <- shiny::reactive({
        shiny::req(se_data(), input$n_top)
        
        run_pca(
            se_data(),
            n_top = input$n_top,
            log_transform = input$log_transform,
            scale = input$scale
        )
    })
    
    # PCA scatter plot
    output$pca_plot <- shiny::renderPlot({
        shiny::req(pca_result(), input$color_by)
        
        shape <- if (is.null(input$shape_by) || input$shape_by == "None") {
            NULL
        } else {
            input$shape_by
        }
        
        plot_pca(
            pca_result(),
            color_by = input$color_by,
            shape_by = shape,
            point_size = input$point_size
        )
    })
    
    # Variance plot
    output$variance_plot <- shiny::renderPlot({
        shiny::req(pca_result())
        
        var_df <- pca_variance_explained(pca_result())
        var_df <- var_df[seq_len(min(8, nrow(var_df))), ]
        var_df$PC <- factor(var_df$PC, levels = var_df$PC)
        
        ggplot2::ggplot(var_df, ggplot2::aes(x = .data$PC, y = .data$variance_percent)) +
            ggplot2::geom_col(fill = "steelblue") +
            ggplot2::geom_text(
                ggplot2::aes(label = sprintf("%.1f%%", .data$variance_percent)),
                vjust = -0.5, size = 4
            ) +
            ggplot2::theme_minimal(base_size = 14) +
            ggplot2::labs(
                x = "Principal Component",
                y = "Variance Explained (%)"
            ) +
            ggplot2::ylim(0, max(var_df$variance_percent) * 1.15)
    })
    
    # Scores table
    output$scores_table <- DT::renderDataTable({
        shiny::req(pca_result())
        DT::datatable(
            pca_result()$scores,
            options = list(pageLength = 10, scrollX = TRUE)
        )
    })
    
    # Download handler
    output$download_plot <- shiny::downloadHandler(
        filename = function() {
            paste0("pca_plot_", Sys.Date(), ".png")
        },
        content = function(file) {
            shape <- if (is.null(input$shape_by) || input$shape_by == "None") {
                NULL
            } else {
                input$shape_by
            }
            
            p <- plot_pca(
                pca_result(),
                color_by = input$color_by,
                shape_by = shape,
                point_size = input$point_size
            )
            
            ggplot2::ggsave(file, p, width = 8, height = 6, dpi = 150)
        }
    )
}
```

### R/run_app.R

```{r run-app}
# R/run_app.R

#' Run the sePCA Shiny Application
#'
#' Launches an interactive Shiny application for exploring PCA results
#' on SummarizedExperiment data.
#'
#' @param ... Additional arguments passed to [shiny::shinyApp()]
#'
#' @return A Shiny app object (invisibly)
#' @export
#'
#' @examples
#' \dontrun{
#' run_app()
#' }
run_app <- function(...) {
    # Check for required packages
    if (!requireNamespace("shiny", quietly = TRUE)) {
        stop("Package 'shiny' is required. Install with: install.packages('shiny')")
    }
    if (!requireNamespace("bslib", quietly = TRUE)) {
        stop("Package 'bslib' is required. Install with: install.packages('bslib')")
    }
    if (!requireNamespace("DT", quietly = TRUE)) {
        stop("Package 'DT' is required. Install with: install.packages('DT')")
    }
    if (!requireNamespace("airway", quietly = TRUE)) {
        stop("Package 'airway' is required. Install with: BiocManager::install('airway')")
    }
    
    app <- shiny::shinyApp(
        ui = app_ui(),
        server = app_server,
        ...
    )
    
    shiny::runApp(app)
}
```

## Step 3: Create inst/app/app.R (Optional Entry Point)

This allows running the app via `shiny::runApp(system.file("app", package = "sePCA"))`:

```{r inst-app}
# inst/app/app.R

# Launch the sePCA Shiny application
# This file exists for compatibility with runApp()
# Preferred method: sePCA::run_app()

if (!requireNamespace("sePCA", quietly = TRUE)) {
    stop("Please install sePCA first: remotes::install_github('you/sePCA')")
}

sePCA:::app_ui
sePCA:::app_server

shiny::shinyApp(
    ui = sePCA:::app_ui(),
    server = sePCA:::app_server
)
```

---

# Part 2: Dependencies for the App

## Updating DESCRIPTION

The Shiny app needs additional packages. Add them as Suggests (not Imports) so the core package works without them:

```{r add-suggests}
library(usethis)

# Shiny app dependencies
use_package("shiny", type = "Suggests")
use_package("bslib", type = "Suggests")
use_package("DT", type = "Suggests")
use_package("airway", type = "Suggests")
use_package("rlang")  # For .data pronoun in ggplot2 aes()
```

Your DESCRIPTION now includes:

```yaml
Suggests:
    airway,
    bslib,
    DT,
    knitr,
    shiny,
    testthat (>= 3.0.0)
```

## Why Suggests?

- Users who only want the R API don't need shiny
- Keeps the core package lightweight
- `run_app()` checks for packages at runtime

## Graceful Fallbacks

What if DT isn't installed? Handle it gracefully:

```{r fallback}
# In app_ui.R, instead of:
DT::dataTableOutput("table")

# Use:
if (requireNamespace("DT", quietly = TRUE)) {
    DT::dataTableOutput("table")
} else {
    shiny::tableOutput("table")  # Basic fallback
}
```

> **Exercise A:** Ensure `run_app()` works after installing from GitHub (no local relative paths). Test in a fresh R session.

---

# Part 3: Documentation

## Document run_app()

We already added roxygen2 documentation. Run:

```{r document}
devtools::document()
```

Verify:

```{r check-docs}
devtools::load_all()
?run_app
```

## Add pkgdown Article

```{r vignette}
usethis::use_vignette("shiny-app", title = "Using the Shiny App")
```

Edit `vignettes/shiny-app.Rmd`:

````markdown
---
title: "Using the Shiny App"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using the Shiny App}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Launching the App

After installing sePCA, launch the interactive PCA explorer:

```r
library(sePCA)
run_app()
```

## Features

### Analysis Settings

- **Top variable genes**: Number of most variable genes to include in PCA
- **Log-transform**: Apply log2(x + 1) transformation to counts
- **Scale features**: Center and scale genes before PCA

### Visualization Options

- **Color by**: Choose a metadata column to color samples
- **Shape by**: Optionally map a second variable to point shape
- **Point size**: Adjust point size for visibility

### Outputs

1. **PCA Plot**: Interactive scatter plot of principal components
2. **Variance**: Bar chart showing variance explained by each PC
3. **Sample Data**: Table of PCA scores merged with sample metadata

### Downloading Results

Click "Download Plot" to save the current PCA plot as a PNG file.
````

> **Exercise B:** Add screenshots to your vignette to make it more visual.

---

# Part 4: Advanced Interactions

## Adding Tabs with Multiple Views

```{r tabs}
bslib::navset_card_tab(
    bslib::nav_panel("PCA Plot", plotOutput("pca_plot")),
    bslib::nav_panel("Loadings", plotOutput("loadings_plot")),
    bslib::nav_panel("Heatmap", plotOutput("heatmap")),
    bslib::nav_panel("Data", DT::dataTableOutput("scores_table"))
)
```

## Interactive Brushing

Let users select points on the plot:

```{r brushing}
# UI
plotOutput("pca_plot", brush = brushOpts(id = "plot_brush"))

# Server
output$selected_samples <- renderTable({
    req(input$plot_brush)
    scores <- pca_result()$scores
    brushedPoints(scores, input$plot_brush, xvar = "PC1", yvar = "PC2")
})
```

## Download Handlers

### Download Plot

```{r download-plot}
output$download_plot <- downloadHandler(
    filename = function() paste0("pca_", Sys.Date(), ".png"),
    content = function(file) {
        ggplot2::ggsave(file, current_plot(), width = 8, height = 6)
    }
)
```

### Download Data (TSV)

```{r download-data}
output$download_scores <- downloadHandler(
    filename = function() paste0("pca_scores_", Sys.Date(), ".tsv"),
    content = function(file) {
        write.table(pca_result()$scores, file, sep = "\t", row.names = FALSE)
    }
)
```

> **Exercise C:** Pair up. Try to crash your partner's app with weird inputs (extreme values, rapid clicking, etc.). Add validation to prevent it.

---

# Part 5: Testing Shiny Apps

## The Challenge

Testing Shiny apps is harder than testing regular functions:

- Apps are stateful (inputs persist)
- Outputs depend on UI interactions
- Need a running R process

## Testing Strategy Options

### 1. Test the Logic, Not the App

The best approach: keep logic in testable functions, test those:

```{r test-logic}
# The core logic is already tested
test_that("run_pca works", { ... })
test_that("plot_pca works", { ... })

# The Shiny app just wires things together
```

### 2. Smoke Tests with shinytest2

For end-to-end testing:

```{r shinytest2}
# Install
install.packages("shinytest2")

# Set up
shinytest2::use_shinytest2()
```

Create `tests/testthat/test-app.R`:

```{r app-test}
test_that("app starts without error", {
    skip_on_cran()  # Skip on CRAN (no display)
    skip_if_not_installed("shinytest2")
    
    app <- shinytest2::AppDriver$new(
        app_dir = system.file("app", package = "sePCA"),
        name = "pca-app"
    )
    
    # Check initial state
    expect_equal(app$get_value(input = "n_top"), 500)
    
    # Change an input
    app$set_inputs(n_top = 1000)
    
    # Wait for outputs to update
    app$wait_for_idle()
    
    # Take a snapshot (optional)
    app$expect_screenshot()
    
    app$stop()
})
```

### 3. Manual Testing Checklist

For HW1, a manual checklist is acceptable:

- [ ] App launches without error
- [ ] Changing n_top updates the plot
- [ ] Changing color_by updates colors
- [ ] Download button produces a file
- [ ] No errors in the R console during use

---

# Part 6: Deployment 

Naturally, we want to share our app with others! There are a few different approaches to do this, but the simplest is to deploy to Posit Cloud for free.

## Option 1: Posit Cloud

1. Create a new project from Git
2. Install your package: `remotes::install_github("you/sePCA")`
3. Create an `app.R` that calls `sePCA::run_app()`
4. Click "Publish"


## Record the URL

After deploying, add to your README:

```markdown
## Interactive App

Try the app online: https://your-account.shinyapps.io/sePCA/

Or run locally:

```r
sePCA::run_app()
```
```

---

# Summary

Today we:

1. Packaged the Shiny app with proper structure (`R/` for logic, `inst/app/` for entry point)
2. Added `run_app()` as a documented, exported function
3. Used Suggests for optional Shiny dependencies
4. Added interactive features (tabs, download handlers)
5. Discussed testing strategies

## Package Milestone

✅ The Shiny app is shipped inside the package and can be launched by users after installing from GitHub.

---

## Debrief & Reflection

Before moving on, make sure you can answer:

- Why is packaging the app part of the reproducibility story, not just a convenience feature?
- Which app dependencies should be mandatory, and which should remain optional with graceful fallback?
- If a collaborator never opens the app and only uses the R API, how does your package design still serve them well?

---

# After-Class Tasks

## Micro-task 1: GitHub Test

From a fresh R session:

```{r fresh-test, eval=FALSE}
# Start clean
.rs.restartR()

# Install from GitHub
remotes::install_github("you/sePCA")

# Run the app
sePCA::run_app()
```

## Micro-task 2: Update README

Add an "Interactive App" section to README with:

- Instructions to launch
- Screenshot or GIF (optional but nice!)
- Link to deployed version (if you deployed)

## Optional: Deploy

Deploy to shinyapps.io and record the URL in your README.

---

# Complete File Reference

## Final Directory Structure

```
sePCA/
├── DESCRIPTION
├── NAMESPACE
├── LICENSE.md
├── README.Rmd
├── README.md
├── R/
│   ├── data.R
│   ├── pca.R
│   ├── plotting.R
│   ├── export.R
│   ├── app_ui.R
│   ├── app_server.R
│   └── run_app.R
├── inst/
│   └── app/
│       └── app.R
├── man/
│   ├── run_app.Rd
│   └── ...
├── tests/
│   ├── testthat/
│   │   └── ...
│   └── testthat.R
├── vignettes/
│   ├── getting-started.Rmd
│   └── shiny-app.Rmd
└── _pkgdown.yml
```

---

# Session Info

```{r session-info, eval=TRUE}
sessionInfo()
```
